```{r include=FALSE}
source("R/common.R")
knitr::opts_chunk$set(fig.path = "figs/ch12/")
```

::: {.content-visible unless-format="pdf"}
{{< include latex/latex-commands.qmd >}}
:::

# Visualizing Equality of Covariance Matrices {#sec-eqcov}

> _To make the preliminary test on variances is rather like putting to sea in a rowing boat
> to find out whether conditions are sufficiently calm for an ocean liner to leave port._
> --- G. E. P. Box [-@Box:1953]


This chapter concerns the extension of tests of homogeneity of variance from the classical univariate
ANOVA setting to the analogous multivariate (MANOVA) setting. Such tests are a routine but important aspect of data analysis, as particular violations can drastically impact model estimates and appropriate conclusions that can be drawn [@Lix:1996]. 

Beyond issues of model assumptions, 
the question of equality of covariance matrices is often of general interest itself. For instance,
variability is often an important issue in studies of strict equivalence in laboratories comparing results across  multiple patient measurements and in other applied contexts [see @Gastwirth-etal:2009 for other exemplars].

Moreover the outcome of such tests often have important consequences
for the details of a main method of analysis.  Just as the Welch $t$-test [@Welch:1947] is now commonly
used and reported for a two-group test of differences in means under _unequal_ variances,
a preliminary test of equality of covariance matrices is often used in discriminant analysis
to decide whether linear (LDA) or quadratic discriminant analysis (QDA) should be applied in a
given problem. In such cases, the data at hand should inform the choice of statistical analysis to utilize.


I provide some answers to the following questions:

- **Visualization**: How can we visualize differences among group variances and covariance matrices,
  perhaps in a way that is analogous to what is done to visualize differences among group means?
  As will be illustrated (@sec-viz-boxM), differences among covariance matrices are comprised of spread in overall _size_ ("scatter") and _shape_ ("orientation"). These can be easily seen in data space with data ellipses, particularly if the data is centered by shifting all groups to the grand mean so they can be directly compared.
  
- **Low-D views**:  When there are more than a few response variables, what low-dimensional
  views can show the most interesting properties related to the equality of covariance matrices?
  Projecting the data into the space of the principal components serves well again here (@sec-eqcov-low-rank-views). Surprisingly, we will see that the _small dimensions_ contain useful information about differences among the group covariance matrices, similar to what we find for outlier detection.
  
- **Other statistics**: Box's $\mathcal{M}$-test [@Box:1949] is most widely used. Are there other worthwhile test statistics? You will see that graphical methods suggest alternatives in @sec-other-measures. Similarly, the univariate
Levene's test has a simple multivariate generalization which can be visualized using HE plots (@sec-multivar-levene).

The following sections provide a capsule summary of the issues in this topic following @FriendlySigal:2018:eqcov.
Most of the discussion is couched in terms of a one-way design for simplicity,
but the same ideas can apply to two-way (and higher) designs, where a "group" factor
is defined as the product combination (interaction) of two or more factor
variables. 

When there are also numeric covariates, this topic can also be extended
to the multivariate analysis of covariance (MANCOVA) setting. This is
accomplished simply by applying these techniques to the _residuals_ from predictions by 
the covariates alone.

**Packages**

In this chapter I use the following packages. Load them now.
```{r pkg-load}
library(car)
library(heplots)
library(candisc)
library(ggplot2)
library(dplyr)
library(tidyr)
```

The main methods described here are implemented in the `r pkg("heplots")` package.

## Homogeneity of Variance in Univariate ANOVA {#sec-homogeneity-ANOVA}

In classical (Gaussian) univariate ANOVA models, the main interest is typically on
tests of mean differences in a response $y$ according to one or more factors.
The validity of the typical $F$ test, however, relies on the assumption
of _homogeneity of variance_: 
all groups have the same (or similar) variance,

$$
\sigma_1^2 = \sigma_2^2 = \cdots = \sigma_g^2 \; .
$$

It turns out that the $F$ test for differences in means
is relatively robust to violation of this assumption [@Harwell:1992],
as long as the group sample sizes are roughly equal.[^liberal]
This applies to Type I error $\alpha$ rates, which are not much affected.
However, unequal variance makes the ANOVA tests less efficient: you lose power to detect
significant differences.

[^liberal]: If group sizes are greatly unequal __and__ homogeneity of variance is violated, then the $F$ statistic is too liberal (actual rejection rate greater than the nominal $\alpha$) when large sample variances are associated with small group sizes.
Conversely, the $F$ statistic is too conservative (actual $< \alpha$) if large variances are associated with large group sizes.

A variety of classical test statistics for homogeneity of variance are available, including
Hartley's $F_{max}$ [@Hartley:1950], Cochran's _C_ [@Cochran:1941],and Bartlett's test [@Bartlett:1937],
but these have been found to have terrible statistical properties [@Rogan:1977], which prompted Box's famous quote.

Levene [-@Levene:1960] introduced a different form of test, based on the simple idea that when 
variances are equal across groups, 
the average _absolute values_ of differences between the observations and group means
will also be equal, i.e., substituting an $L_1$ norm for the $L_2$ norm of variance. 
In a one-way design, this is equivalent to a test of group differences in the
means of the auxilliary variable $z_{ij} = | y_{ij} - \bar{y}_i |$.



More robust versions of this test were proposed by @BrownForsythe:1974. These tests
substitute the group mean by either the group **median** or a **trimmed mean** in the ANOVA of the absolute deviations.
Some suggest these should be almost always preferred to Levene's version using the mean deviation.
See @Conover-etal:1981 for an early review and @Gastwirth-etal:2009 for a general discussion of these tests. 
In what follows, we refer to this class of tests as "Levene-type" tests and suggest a 
multivariate extension described below (@sec-mlevene).

These deviations from a group central can be calculated using `heplots::colDevs()` and the central value can be
a function, like `mean`, `median` or an anonymous one like `function(x) mean(x, trim = 0.1))` that trims 10%
off each side of the distribution. With a response `Y` Levene-type tests then be performed "by hand" as follows:

```{r lm-levene, eval=FALSE}
# Levene
Z.mean <- abs( colDevs(Y, group) )
lm(Z.mean ~ group)

# Brown-Forsythe
Z.med <- abs( colDevs(Y, group, median) )
lm(Z.med ~ group)
```

The function `car::leveneTest()` does this, so we could examine whether the variances are equal in the Penguin variables, one at a time,
like so:
```{r levene-peng1}
data(peng, package = "heplots")
leveneTest(bill_length ~ species, data=peng)

leveneTest(bill_depth ~ species, data=peng)

# ...

leveneTest(body_mass ~ species, data=peng)

```


More conveniently, `heplots:leveneTests()` with an "s", does this for each of a set of response variables,
specified in a data frame, a model formula or a `"mlm"` object.
It also formats the results in a more pleasing way:

```{r levene-peng2}
peng.mod <- lm(cbind(bill_length, bill_depth, flipper_length, body_mass) ~ species, 
               data = peng)
leveneTests(peng.mod)
```

So, this tells us that the groups do not differ in variances on first three variables, but they do for `body_mass`.

## Visualizing Levene's test {#sec-mlevene}

To gain some insight into the problem of homogeneity of variance it is helpful so see how the situation looks in terms of data.
For the Penguin data, it might be simplest just just to look at boxplots of the variables and try to see whether the 
**widths** of the
central 50% boxes seem to be the same, as in @fig-peng-boxplots.
However, it is perceptually difficult to focus on differences with widths of
the boxes within each panel when their centers also differ from group to group.

```{r}
#| label: fig-peng-boxplots
#| code-fold: true
#| code-summary: "See the code"
#| fig-width: 10
#| fig-height: 5
#| out-width: "80%"
#| fig-cap: "Boxplots for the Penguin variables. For assessing homogeneity of variance, we should be looking for differences in **width** of
#| the central 50% boxes in each panel, rather than difference in central tendency."
source("R/penguin/penguin-colors.R")
col <- peng.colors("dark")
clr <- c(col, gray(.20))
peng_long <- peng |> 
  pivot_longer(bill_length:body_mass, 
               names_to = "variable", 
               values_to = "value") 

peng_long |>
  group_by(species) |> 
  ggplot(aes(value, species, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = 'free_x') +
  theme_penguins() +
  theme_bw(base_size = 14) +
  theme(legend.position = 'none') 
```

Instead, you can see more _directly_ what is tested by the Levene test by graphing the _absolute deviations_ from the group means or medians. This is another example of the graphic idea that you can make visual comparisons easier
by plotting quantities of direct interest.
You can calculate the median deviation values as follows:

```{r}
vars <- c("bill_length", "bill_depth", "flipper_length", "body_mass")
pengDevs <- colDevs(peng[, vars], peng$species, median) |>
  abs()
```

From a boxplot of the absolute deviations in @fig-peng-devplots your eye can now focus on the central
value, shown by the **median '|'** line, because Levene's method is testing whether these differ
across groups.

```{r}
#| label: fig-peng-devplots
#| code-fold: true
#| code-summary: "See the code"
#| fig-width: 10
#| fig-height: 5
#| out-width: "80%"
#| fig-cap: "Boxplots for absolute differences from group medians for the Penguin data. The visual test of equality of variance is whether the median lines in the boxplots align." 
# calculate absolute differences from median
dev_long <- data.frame(species = peng$species, pengDevs) |> 
  pivot_longer(bill_length:body_mass, 
               names_to = "variable", 
               values_to = "value") 

dev_long |>
  group_by(species) |> 
  ggplot(aes(value, species, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = 'free_x') +
  xlab("absolute median deviation") +
  theme_penguins() +
  theme_bw(base_size = 14) +
  theme(legend.position = 'none') 
```

It is now easy to see that the medians largely align for all the variables except for `body_mass`.

## Homogeneity of variance in MANOVA {#sec-homogeneity-MANOVA}

In the MANOVA context, the main emphasis, of course, is on differences among mean vectors,
testing

$$
\mathcal{H}_0 : \boldsymbol{\mu}_1 = \boldsymbol{\mu}_2 = \cdots = \boldsymbol{\mu}_g \; .
$$
However, the standard test statistics (Wilks' Lambda, Hotelling-Lawley trace, Pillai-Bartlett trace, Roy's maximum root) rely upon
the analogous assumption that
the within-group covariance **matrices** $\boldsymbol{\Sigma}_i$ are equal for all groups,

$$
\boldsymbol{\Sigma}_1 = \boldsymbol{\Sigma}_2 = \cdots = \boldsymbol{\Sigma}_g \; .
$$
This is much stronger than in the univariate case, because it also requires that all the correlations between pairs of variables are the same for all groups. For example, for two responses, there are three
parameters ($\rho, \sigma_1^2, \sigma_2^2$) assumed equal across all groups; for $p$ responses, there
are $p (p+1) / 2$ assumed equal. The variances relate to _size_ differences among data ellipses
while the differences in the covariances appear as differences in _shape_.

<!--
For example, in the case of two responses,
we must assume:

$$
\begin{pmatrix} 
\sigma_1^2             & \textsf{sym} \\ 
\rho \sigma_1 \sigma_2 & \sigma_2^2   \\ 
\end{pmatrix}_1 = 
\begin{pmatrix} 
\sigma_1^2             & \textsf{sym} \\ 
\rho \sigma_1 \sigma_2 & \sigma_2^2   \\ 
\end{pmatrix}_2 = \dots =
\begin{pmatrix} 
\sigma_1^2             & \textsf{sym} \\ 
\rho \sigma_1 \sigma_2 & \sigma_2^2   \\ 
\end{pmatrix}
_g
$$
-->

::: {#exm-peng-covellipses}
**Penguin data**: Covariance ellipses

\ixon{peng data}
\ixon{datasets!peng}

To preview a main example, @fig-peng-covEllipse0 shows data ellipses for the
main size variables in the Penguins data (`peng`).
<!-- \ix{datasets!peng} -->
<!-- \ix(peng data) -->
`heplots::covEllipses()` is specialized for viewing the relations among 
the data ellipsoids representing the sample covariance matrices,
$\mathbf{S}_1 = \mathbf{S}_2 = \cdots = \mathbf{S}_g$.
It draws the data ellipse for each group, and also for the pooled within-group
$\mathbf{S}_p$, as shown in @fig-peng-covEllipse0 for bill length and bill depth.

You can see that the sizes and shapes of the data ellipses are sort of similar
in the left panel. The visual comparison becomes more precise when the data
ellipses are all shifted to a common origin at the grand means (using `center = TRUE`).
From this you can see that the Adélie group differs most from the others.

```{r}
#| label: fig-peng-covEllipse0
#| fig-align: center
#| fig-height: 5
#| fig-width: 10
#| out-width: "90%"
#| code-fold: true
#| code-summary: "See the code"
#| fig-cap: "Data ellipses for bill length and bill depth in the penguins data, also showing the pooled covariance. Left: As is; right: these are centered at the grand means for easier comparison."
op <- par(mar = c(4, 4, 1, 1) + .5,
          mfrow = c(c(1,2)))
covEllipses(cbind(bill_length, bill_depth) ~ species, data=peng,
  fill = TRUE,
  fill.alpha = 0.1,
  lwd = 3,
  col = clr)

covEllipses(cbind(bill_length, bill_depth) ~ species, data=peng,
  center = TRUE, 
  fill = c(rep(FALSE,3), TRUE), 
  fill.alpha = .1, 
  lwd = 3,
  col = clr,
  label.pos = c(1:3,0))
par(op)
```

All such pairwise plots in scatterplot matrix format are produced using the `variables` argument to `covEllipses()`, giving
@fig-peng-covEllipse-pairs.
```{r}
#| label: fig-peng-covEllipse-pairs
#| fig-align: center
#| fig-width: 8
#| fig-height: 8
#| out-width: "90%"
#| code-fold: true
#| code-summary: "See the code"
#| fig-cap: "All pairwise covariance ellipses for the penguins data. The covariance matrices are homogeneous when the ellipses for the groups all have the same size and shape as that for the mean-centered pooled data (shown in black)."
clr <- c(peng.colors(), "black")
covEllipses(peng[,3:6], peng$species, 
  variables=1:4,
  col = clr,
  fill=TRUE, 
  fill.alpha=.1)
```

The covariance ellipses in @fig-peng-covEllipse-pairs look pretty similar in size, shape and orientation.
But what does Box's $\mathcal{M}$ test (described below) say? 

```{r boxm}
peng.boxm <- boxM(cbind(bill_length, bill_depth, flipper_length, body_mass) 
                  ~ species,  data=peng) |>
  print()
```

As you can see,
it concludes strongly against the null hypothesis, because the test is highly
sensitive to small differences among the covariance matrices.

\ixoff{peng data}
\ixoff{datasets!peng}

:::

::: {#exm-iris-covellipses}
**Iris data**: Covariance ellipses

It will be useful to have another example as we proceed, so @fig-iris-covEllipse-pairs shows an analogous plot for the iris data we examined in @sec-HEplot-matrices. 

Even when these are shown uncentered,
the differences in size, shape and orientation are much more apparent. Iris _setosa_ stands out as
having smaller variance on some of the variables, while the ellipses for _virginica_ tend to be larger. Their orientation (slopes) also differ quite a bit.

```{r}
#| label: fig-iris-covEllipse-pairs
#| fig-align: center
#| fig-width: 8
#| fig-height: 8
#| out-width: "90%"
#| code-fold: show
#| fig-cap: "All pairwise covariance ellipses for the iris data. The large differences in size and shape indicate substantial heterogeneity in variances and covariances."
iris.colors <- c("red", "darkgreen", "blue")
covEllipses(iris[,1:4], iris$Species, 
  variables=1:4, 
  fill = TRUE,
	fill.alpha=.1,
	col = c(iris.colors, "black"),
	label.pos=c(1:3,0))
```

`boxM()` gives the following results:
```{r iris-boxm}
iris.boxm <- boxM(cbind(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) 
                  ~ Species, 
                  data=iris) |>
  print()
```

Unsurprisingly, Box's $\mathcal{M}$ tests gives $\chi^2_{20} = 140.94$ with a $p$-value < 2.2e-16,
putting numbers to the visual conclusion from @fig-iris-covEllipse-pairs.
:::

## Box's $\mathcal{M}$ test {#sec-boxM}

Take a moment and think, "How could we generalize a test of equality of variances, 
$s_1^2 = s_2^2 = \cdots = s_g^2$, to the multivariate case, where we have $(p \times p)$ matrices,
$\mathbf{S}_1 = \mathbf{S}_2 = \cdots = \mathbf{S}_g$ for each of $g$ groups?".

Multivariate thinking suggests that that we calculate some measure of "size" of each $\mathbf{S}_i$,
in a similar way to what is done in multivariate tests comparing $\mathbf{H}$ and $\mathbf{E}$ matrices.

@Box:1949 proposed the following likelihood-ratio test (LRT) statistic $\mathcal{M}$ for testing the
hypothesis of equal covariance matrices. It uses the log of the determinant $\vert \mathbf{S}_i \vert$
as the measure of size.

$$
\mathcal{M} = (N -g) \; \ln \vert \mathbf{S}_p \vert - \sum_{i=1}^g (n_i -1) \; \ln \vert \mathbf{S}_i \vert \; ,
$$ {#eq-boxm}

where $N = \sum n_i$ is the total sample size and

$$
\mathbf{S}_p = (N-g)^{-1} \sum_{i=1}^g (n_i - 1) \mathbf{S}_i
$$

is the pooled covariance matrix.

$\mathcal{M}$ can thus be thought of as a ratio of the determinant of the pooled $\mathbf{S}_p$
to the geometric mean of the determinants of the separate $\mathbf{S}_i$.

In practice, there are various transformations of the value of $M$ to yield a
test statistic with an approximately known distribution [@Timm:75].
Roughly speaking, when each $n_i > 20$, a $\chi^2$ approximation is often used;
otherwise an $F$ approximation is known to be more accurate.

Asymptotically, $-2 \ln (\mathcal{M})$ has a $\chi^2$ distribution. The $\chi^2$ approximation
due to Box [-@Box:1949; -@Box:1950] is that

$$
X^2 = -2 (1-c_1) \; \ln (\mathcal{M}) \quad \sim \quad \chi^2_\text{df}
$$
with $\text{df} = (g-1) p (p+1)/2$ degrees of freedom. A bias correction constant $c_1$ improves accuracy of the
approximation:

$$
c_1 = \left( 
\sum_i \frac{1}{n_i -1}
- \frac{1}{N-g}
\right)
\frac{2p^2 +3p -1}{6 (p+1)(g-1)} \; .
$$

In this form, Bartlett's test for equality of variances in the univariate case
is the special case of Box's $\mathcal{M}$ when there is only one response variable,
so Bartlett's test is sometimes used as univariate follow-up to determine
which response variables show heterogeneity of variance.

Yet, like its univariate counterparts, Box's test is well-known to be highly sensitive to violation
of (multivariate) normality and the presence of outliers[^boxm-problems], as @Box:1953 suggested
in the opening chapter quote.
Yet, it provides a nice framework for thinking about this problem more generally and gives rise
to interesting graphic methods.

[^boxm-problems]: For example, @TikuBalakrishnan:1984
concluded from simulation studies that the normal-theory LRT provides poor control of Type I
error under even modest departures from normality. @OBrien:1992 proposed some robust alternatives,
and showed that Box's normal theory approximation suffered both in controlling the null size
of the test and in power. @ZhangBoos:1992:BCV also carried out simulation studies with similar
conclusions and used bootstrap methods to obtain corrected critical values.

## Visualizing heterogeneity

A larger goal of this chapter is to use this background as another illustration of multivariate thinking,
here, for visualizing and testing the heterogeneity of covariance matrices in multivariate designs.  

While researchers often rely on a single number to determine if their data have met a particular threshold, such compression will often obscure interesting information, particularly when a test concludes that differences exist, and one is left to wonder ``why?''. It is within this context where, again, visualizations often reign supreme. 

We have already seen one useful method in @sec-homogeneity-MANOVA, which uses
direct visualization of the information in the $\mathbf{S}_i$ and $\mathbf{S}_p$ using _data ellipsoids_ to show size and shape as minimal schematic summaries;
In what follows, I propose three additional visualization-based approaches to questions of heterogeneity of covariance in MANOVA
designs:

* a simple **dotplot** (@sec-viz-boxM) of the components of Box's $\mathcal{M}$ test: the log determinants of the 
$\mathbf{S}_i$ together with that of the pooled $\mathbf{S}_p$. Extensions of these simple plots raise the question of whether measures of heterogeneity other than that captured in Box's test might also be useful; and,

* PCA **low-rank views** (@sec-eqcov-low-rank-views) to highlight features more easily seen there than in the full data space.

* The analogy with the four types of multivariate test statistics (@sec-multivar-tests) suggests other measures
that could be used to compare the $\mathbf{S}_i$ with $\mathbf{S}_p$ (@sec-other-measures).

* the connection between Levene-type tests and an ANOVA (of centered absolute differences) suggests a parallel with a multivariate extension of Levene-type tests and a MANOVA (@sec-multivar-levene).  We explore this with a version of
Hypothesis-Error (HE) plots that we found useful for visualizing mean differences in MANOVA designs.

These methods are described and illustrated in @FriendlySigal:2018:eqcov.

## Visualizing Box's $\mathcal{M}$ {#sec-viz-boxM}

Box's test is based on a comparison of the log determinants of the $\mathbf{S}_i$ relative to that of the pooled $\mathbf{S}_p$, so the simplest thing to do is just plot them!

`boxM()` produces a `"boxm"` object, for which there are `summary()` (details) and `plot()` methods.
The `plot()` method gives a dot plot of the log determinants $\ln \vert \mathbf{S}_i \vert$
together with that for the pooled covariance $\ln \vert \mathbf{S}_p \vert$.
@Cai-etal:2015 provide the theory for the (asymptotic) confidence intervals shown.

```{r echo = -1}
#| label: fig-peng-iris-boxm-plots
#| fig-width: 10
#| fig-height: 5
#| out-width: "90%"
#| fig-cap: "Plots of the contributions to Box's $\\mathcal{M}$ statistic for the Penguin and iris data."
op <- par(mar=c(4,5,1,1)+.1, mfrow = c(1,2))
plot(peng.boxm, gplabel="species", cex.lab = 1.5)

plot(iris.boxm, gplabel="Species", cex.lab = 1.5)
```

In these plots (@fig-peng-iris-boxm-plots), the value for the pooled covariance appears within the range of the groups, because it is a weighted average.
If you take a moment to look back at @fig-peng-covEllipse-pairs, you'll see that the data ellipses
for Gentoo are slightly smaller in most pairwise views, but it is much easier to see
this in a plot that summarizes this, like @fig-peng-iris-boxm-plots.

From @fig-iris-covEllipse-pairs,
it is clear that _setosa_ shows the smallest within-group variability.
The numeric scale values on the horizontal axis give a sense that the range across groups is
considerably greater for the iris data than for the Penguins.
Some generalizations of Box's test using other measures are illustrated
in @sec-other-measures.

<!--
Just as multivariate tests provide different statistics (Wilks' $\Lambda$, Pillai trace, ...) based on different functions of eigenvalues (@tbl-mstats), the idea of Box's $\mathcal{M}$ suggests other possible test statistics to measure the relative sizes
of the covariance matrices $\mathbf{S}_i$ and $\mathbf{S}_p$.
For example, $\text{tr}(S) = \Sigma_j \lambda_j$ is measure of average variance
based on the sum over dimensions and $1/ \text{tr}(S^{-1}) = 1/ \Sigma_j (1/\lambda_j) $
is a measure of average precision, corresponding to a harmonic mean.

The `plot.boxm()` method can compute and plot these and other measures,
controled by the `which =` argument.
@fig-peng-iris-boxm-plots2 shows plots of the sum and precision statistics for
the iris data.

```{r echo=-1}
#| label: fig-peng-iris-boxm-plots2
#| fig-width: 10
#| fig-height: 5
#| out-width: "100%"
#| fig-cap: "Plots of the other eigenvalue statistics for the iris data."
op <- par(mfrow=c(1,2), mar=c(5,4,1,1))
plot(iris.boxm, which="sum", gplabel="Species")
plot(iris.boxm, which="precision", gplabel="Species")
par(op)
```
-->


## Low-rank views {#sec-eqcov-low-rank-views}

With $p>3$ response variables, a simple alternative to the pairwise 2D plots in data space (shown in 
@fig-peng-covEllipse-pairs and @fig-iris-covEllipse-pairs) is the projection into the principal component space accounting for the
greatest amounts of total variance in the data (@FriendlySigal:2018:eqcov).
For the Iris data, a simple PCA of the covariance matrix shows that nearly 96\% of total variance in the data
is accounted for in the first two dimensions.

```{r}
iris.pca <- prcomp(iris[,1:4], scale. = TRUE)
summary(iris.pca)
```


@fig-iris-pca-covellipses shows the plots of the covariance ellipsoids for the first two principal component scores,
uncentered (left panel) and centered (right panel). The dominant PC1 (73\% of total variance)
essentially orders the species by a measure of overall size of their sepals and petals.
In the centered view at the right, it can again be seen how _Setosa_ differs in
covariance from the other two species, and that while _Virginca_ and _Versicolor_ both have similar
shapes to the pooled covariance matrix, _Versicolor_ has somewhat greater variance on PC1.

```{r echo=-1}
#| label: fig-iris-pca-covellipses
#| fig-height: 6
#| fig-width: 12
#| out-width: "90%"
#| fig-cap: "Covariance ellipsoids for the first two principal components of the iris data. Left (a): uncentered, showing group means on the principal components; right (b): centered at the origin, making size and shape comparison easier."
op <- par(mfcol=c(1,2), mar=c(5,4,1,1)+.1)
source(here::here("R/util/text.usr.R"))    # text in user coords

covEllipses(iris.pca$x, iris$Species, 
  col = c(iris.colors, "black"),                # uncentered
	fill=TRUE, fill.alpha=.1,
  cex.lab = 1.5,
	label.pos = c(1, 3, 3, 0), asp=1)
text.usr(0.05, 0.95, "(a) Uncentered", pos = 4, cex = 1.5)

covEllipses(iris.pca$x, iris$Species,
  center=TRUE,                                  # centered
  col = c(iris.colors, "black"),
  fill=TRUE, fill.alpha=.1,
  cex.lab = 1.5,
  label.pos = c(1, 3, 3, 0), asp=1)
text.usr(0.95, 0.95, "(b) Centered", pos = 2, cex = 1.5)
```

### Small dimensions can matter

For the Iris data, the first two principal components account for 96\% of total variance,
so we might think we are done here. Yet, as we've seen in other problems
(outliers, collinearity), important information also exists in the space of the _smallest_ principal component dimensions.

This is also true, as we will see for Box's $\mathcal{M}$ test, because it is a (linear) function of all the eigenvalues
of the between and within group covariance matrices, is therefore also subject to the influence of the smaller dimensions, where differences among $\mathbf{S}_i$ and of $\mathbf{S}_p$ can lurk.

```{r echo=-1}
#| label: fig-iris-pca-covellipses-dim34
#| fig-height: 6
#| fig-width: 6
#| out-width: "60%"
#| fig-cap: "Covariance ellipses for the smallest principal components of the iris data."
op <- par(mar=c(5,4,1,1)+.1)
covEllipses(iris.pca$x, iris$Species,
  variables = 3:4,
  center=TRUE,        
  col = c(iris.colors, "black"),
  fill=TRUE, fill.alpha=.1,
  cex.lab = 1.5,
  label.pos = c(1, 3, 3, 4), asp=1)
```

@fig-iris-pca-covellipses-dim34 shows the covariance ellipsoids in (PC3, PC4) space. 
Even though these dimensions contribute little to total variance, there are more 
pronounced differences in the within-group shapes (correlations) relative to
the pooled covariance, and these contribute to a rejection of homogeneity by Box's $\mathcal{M}$ test.
Here we see that the correlation for _Virginca_ is of opposite sign from the
other two groups.

## Other measures of heterogeneity {#sec-other-measures}

As we saw above (@sec-homogeneity-MANOVA), the question of equality of covariance matrices
can be expressed in terms of the similarity in size and shape of the data ellipses
for the individual group $\mathbf{S}_i$ relative to that of $\mathbf{S}_p$.
Box's $\mathcal{M}$ test uses just one possible function to describe this size: the logs of their determinants.

When $\boldsymbol{\Sigma}$ is the covariance matrix of a multivariate vector $\vec{y}$ with eigenvalues
$\lambda_1 \ge \lambda_2 \ge \dots \lambda_p$,
the properties shown in @tbl-eigval-ellipse represent methods of describing the size and shape of the ellipsoid in $\Real{p}$.[^geom-ellipsoids]
Just as is the case for tests of the MLM itself
where different functions of these give
test statistics (Wilks' $\Lambda$, Pillai trace, etc.), one could construct
other test statistics for homogeneity of covariance matrices as shown in @tbl-eigval-ellipse.

[^geom-ellipsoids]: More general theory and statistical applications of the geometry of ellispoids is given by
@Friendly-etal:ellipses:2013.

::: {#tbl-eigval-ellipse tbl-pos="!htb"}
```{=latex}
\begin{tabular}{llll}
\textbf{Size}              &  \textbf{Conceptual formula}         & \textbf{Geometry} & \textbf{Function} \\
\hline
(a) Generalized variance:  & $\det{\boldsymbol{\Sigma}} = \prod_i \lambda_i$ & area, (hyper)volume & geometric mean\\  
(b) Average variance:        & $\text{tr}({\boldsymbol{\Sigma}}) = \sum_i \lambda_i $ & linear sum & arithmetic mean\\     
(c) Average precision:        & $1/ \text{tr}({\boldsymbol{\Sigma}^{-1}}) = 1/\sum_i (1/\lambda_i) $ &  & harmonic mean\\
(d) Maximal variance:      & $\lambda_1$ & maximum dimension & supremum
\hline
\end{tabular}
```

Statistical and geometrical properties of "size" of an ellipsoid
:::

Hence, for a sample covariance matrix $\mathbf{S}$, $\vert \mathbf{S} \vert$ is a measure of generalized variance and
$\ln \vert \mathbf{S} \vert$ is a measure of average variance across the $p$ dimensions.

The  plot methods in for `"boxM"` objects in `r pkg("heplots")` can compute and plot all of the functions of the eigenvalues 
in @tbl-eigval-ellipse.  The results for the Penguin data are shown in @fig-peng-boxm-plots.

```{r echo=-1}
#| label: fig-peng-boxm-plots
#| #| fig-show: "hold"
#| fig-height: 5
#| fig-width: 6
#| out-width: "90%"
#| fig-cap: "Plot of eigenvalue statistics of the covariance matrices for the Penguin data."
op <- par(mfrow=c(2,2), mar=c(5,4,1,1), cex.lab = 1.5)
plot(peng.boxm, which="product",   gplabel="species")
plot(peng.boxm, which="sum",       gplabel="species")
plot(peng.boxm, which="precision", gplabel="species")
plot(peng.boxm, which="max",       gplabel="species")
```

Except for the absence of error bars, the plot for log product in the upper left panel of
@fig-peng-boxm-plots is the same as that in @fig-peng-iris-boxm-plots, but with the sign reversed.
In principle, it is possible to add such confidence intervals for all these measures through the use of bootstrapping, but this has not yet been implemented in the `r package("heplots")`.

For this data set, the pattern of points in the plot for Box's $\mathcal{M}$ is also more or less the same 
as that for the precision measure. The plots for the sum of and maximum eigenvalue are also similar to each other, but differ from those of the two measures in the left column of @fig-peng-boxm-plots. 

The main point is that these are not _all the same_, so different functions reflect different patterns of the eigenvalues, and could be used to define new statistical tests, perhaps with greater power or sensitivity to outliers.

## Multivariate analog of Levene's test {#sec-multivar-levene}

The fact that Levene's test is just a simple ANOVA of the absolute deviations
from the group means or medians suggests an easy multivariate generalization
(which has not been well-studied) ---Simply do a MANOVA of the absolute differences,
$\vert \mathbf{Y} - \overline{\mathbf{Y}}_j \vert$, centering on the means or medians.[^Anderson]
For the `iris` data, this gives:

[^Anderson]: @Anderson2006 describes other procedures based on Euclidean _distances_ 
$D_{ij} = [\Sigma (y_{ij} - \bar{y}_j)^2]^{1/2}$
of the observations from their
group centroids (means or spatial medians). Rather than fit a parametric MANOVA model, he suggests a
permutation test method (PERMDIST) which randomly permutes the group assignments and yields non-parametric tests.


```{r iris-dev}
irisdev <- colDevs(iris[,1:4], iris$Species, median, 
                   group.var = "Species") |>
  mutate(across(where(is.numeric), abs))

irisdev.mod <- lm(cbind(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) ~ 
                    Species, data=irisdev)
Anova(irisdev.mod)
```

The `irisdev.mod` model is just another MANOVA, so we can visualize it in an HE plot (@fig-iris-dev-pairs).
But remember that this is showing differences among groups in multivariate dispersion rather
than in means.

```{r}
#| label: fig-iris-dev-pairs
#| fig-width: 8
#| fig.height: 8
#| out-width: "90%"
#| fig-cap: "HE plots for the multivariate generalization of Levene's test based on MANOVA of absolute deviations from group medians."
pairs(irisdev.mod, variables=1:4, fill=TRUE, fill.alpha=.1)
```

In this plot, the $\mathbf{H}$ ellipses for the groups show the variation in the overall _dispersion_---the size of absolute differences
from the group medians. These are very highly correlated across groups in most of the subplots, particularly so
for those involving sepal width.

### Canonical discriminant analysis

But we can go further. Just as canonical discriminant analysis provides a low-dimensional view of differences
in means, the same method can be used to visualize heterogeneity of dispersion expressed by
the the Levene-type MANOVA in 1D or 2D. 

For the model `irisdev.mod`, this shows that 98% of groups differences
shown in the HE plots (@fig-iris-dev-pairs) can be found in a single canonical dimension. The 4D information
can be reduced to 1D.

```{r}
library(candisc)
irisdev.can <- candisc(irisdev.mod) |>
  print()
```

Plotting the first canonical dimension in @fig-iris-dev-can gives a 1D summary, using
boxplots for the scores and vectors for the weights of the iris variables. 

```{r echo=-1}
#| label: fig-iris-dev-can
#| fig-width: 10
#| fig.height: 6
#| out-width: "90%"
#| fig-cap: "Canonical discriminant plot for the multivariate generalization of Levene's test based on MANOVA of absolute deviations from group medians."
op <- par(mar = c(5, 5, 2, 1) + .1)
plot(irisdev.can, which=1,
     fill.alpha = .3,
     col = iris.colors, lwd = 2,
     var.cex = 1.2,
     cex.lab = 1.5, cex.axis = 1.1)
```

Overall, the covariance matrices are smallest for setosa and largest for virginica.
The structure coefficents for the variables reflect the pattern shown in the HE plots (@fig-iris-dev-pairs),
where the direction of the $\mathbf{H}$ ellipses is negative for sepal width but positive for the others.


```{r child = "summary/Ch12-summary.qmd"}
```

<!--
**Packages used here**
-->

```{r write-pkgs}
#| echo: false
#cat("Writing packages to ", .pkg_file, "\n")
write_pkgs(file = .pkg_file, quiet = TRUE)
```


<!-- ## References -->
