```{r include=FALSE}
source("R/common.R")
knitr::opts_chunk$set(fig.path = "figs/case-studies/")
```

::: {.content-visible unless-format="pdf"}
{{< include latex/latex-commands.qmd >}}
:::

# Case Studies

<!--
This chapter presents some complete analyses of datasets that will be prominent in the book.
Some of this material may later be moved to earlier chapters.
-->

::: {.callout-note}
This chapter may be moved from the printed PDF copy to an online Appendix.
:::


For ease of exposition, the chapters in Part IV of this book used small examples to illustrate
the statistical and visualization aspects of the techniques as they were described.
Here I present more complete analyses of a few datasets to illustrate ...

**Packages**

In this chapter I use the following packages. Load them now.
```{r load-pkgs}
library(car)
library(heplots)
library(candisc)
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrgram)
library(ggbiplot)
```

<!-- was: {r child="child/neuro.qmd"} -->

## Neuro- and Social-cognitive measures in psychiatric groups

A Ph.D. dissertation by Laura Hartman [-@Hartman:2016] at York University was designed to evaluate whether and
how clinical patients diagnosed (on the DSM-IV) as schizophrenic or with schizoaffective disorder could be
distinguished from each other and from a normal, control sample on collections of standardized
tests in the following domains:

* **Neuro-cognitive**: processing speed, attention, verbal learning, visual learning and problem solving;

* **Social-cognitive**: managing emotions, theory of mind, externalizing, personalizing bias.

The study is an important contribution to clinical research because the two diagnostic categories
are subtly different and their symptoms often overlap. Yet, they’re very different and often require different treatments. A key difference between schizoaffective disorder and schizophrenia is the prominence of mood disorder involving bipolar, manic and depressive moods.
With schizoaffective disorder, mood disorders are front and center. With schizophrenia, that is not a dominant part of the disorder, but psychotic ideation (hearing voices, seeing imaginary people) is.

### Research questions

This example is concerned with the following substantitive questions:

* To what extent can patients diagnosed as schizophrenic or with schizoaffective disorder  be distinguished from each other and from a normal control sample using a well-validated, comprehensive neurocognitive battery specifically designed for individuals with psychosis  [@Heinrichs-etal:2015]?

* If the groups differ, do any of the cognitive domains show particularly larger or smaller differences among these groups? 

* Do the neurocognitive measures discriminate among the in the same or different ways?
If different, how many separate aspects or dimensions are distinguished?

Apart from the research interest, it could aid diagnosis and treatment if these similar mental disorders could be distinguished by tests in the cognitive domain.


### Data

The clinical sample comprised 116 male and female patients who met the following criteria: 1) a diagnosis of schizophrenia ($n$ = 70) or schizoaffective disorder ($n$ = 46) confirmed by the Structured Clinical Interview for DSM-IV-TR Axis I Disorders; 2) were outpatients; 3) a history free of developmental or learning disability; 4) age 18-65; 5) a history free of neurological or endocrine disorder; and 6) no concurrent diagnosis of substance use disorder.
Non-psychiatric control participants ($n$ = 146) were screened for medical and psychiatric illness and history of substance abuse and were recruited from three outpatient clinics. The dataset is `heplots::NeuroCog`.

```{r NC-data}
data(NeuroCog, package="heplots")
str(NeuroCog)
```

The diagnostic classification variable is called `Dx` in the dataset. To facilitate answering questions regarding group differences, the following contrasts were applied:
the first column compares the control group to the average of the diagnosed groups, the second
compares the schizophrenia group against the schizoaffective group.

```{r NC-contrasts}
contrasts(NeuroCog$Dx)
```


In this analysis, I ignore the `SocialCog` variable because aspects of social cognition were measured more extensively in a separate dataset discussed below (@sec-social-cog).
The primary focus here is on the
the cognitive and perceptual variables `Attention : ProbSolv`.

### A first look

As always, plot the data first! We want an overview of the distributions of the variables to see
the centers, spread,  shape and possible outliers for each group on each variable.

The plot below combines the use of boxplots and violin plots to give an informative display.
As we saw earlier (e.g., @sec-NLSY-mmra), doing this with `ggplot2` requires reshaping the data to
long format.

```{r NC-long}
# Reshape from wide to long
NC_long <- NeuroCog |>
  dplyr::select(-SocialCog, -Age, -Sex) |>
  tidyr::gather(key = response, value = "value", Speed:ProbSolv)
# view 3 observations per group and measure
NC_long |>
  group_by(Dx) |>
  sample_n(3) |> ungroup()
```

In the plot, we take care to adjust the transparency (`alpha`) values for the points,
violin plots and boxplots so that all can be seen. Options for `geom_boxplot()`
are used to give these greater visual prominence.

```{r}
#| label: fig-NC-boxplot
#| fig-align: center
#| out-width: "120%"
#| code-fold: show
#| fig-cap: "Boxplots combined with violin plots of the `NeuroCog` data."
ggplot(NC_long, aes(x=Dx, y=value, fill=Dx)) +
  geom_jitter(shape=16, alpha=0.8, size=1, width=0.2) +
  geom_violin(alpha = 0.1) +
  geom_boxplot(width=0.5, alpha=0.4, 
               outlier.alpha=1, 
               outlier.size = 3, outlier.color = "red") +
  scale_x_discrete(labels = c("Schizo", "SchizAff", "Control")) +
  facet_wrap(~response, scales = "free_y", as.table = FALSE) +
  theme_bw() +
  theme(legend.position="bottom",
        axis.title = element_text(size = rel(1.2)),
        axis.text  = element_text(face = "bold"),
        strip.text = element_text(size = rel(1.2)))
```

We can see that the control participants score higher on all measures, but there is no consistent
pattern of medians for the two patient groups. But these univariate summaries do not inform
about the relations among variables.

## Bivariate views

### Corrgram {.unnumbered}

<!--
A corrgram [@Friendly:02:corrgram] provides a useful reconnaissance plot of the bivariate correlations in the dataset. It suppresses details, and allows focus on the overall pattern.
The `corrgram::corrgram()` function has the ability to enhance perception by
permuting the variables in the order of their variable vectors in a biplot, so more highly correlated variables are adjacent in the plot, and example of _effect ordering_ for
data displays [@FriendlyKwan:03:effect].

The plot below includes all variables except for `Dx` group.
There are a number of `panel.*` functions for choosing how the correlation for each pair is
rendered.
-->

With this many variables, it is useful to view the correlations among them in a corrgram (@sec-corrgram),
giving a high-level reconnaissance.
As noted earlier, the `corrgram::corrgram()` function can enhance perception by
permuting the variables in the order of their variable vectors in a biplot, so more highly correlated variables are adjacent in the plot, an example of _effect ordering_ for
data displays [@FriendlyKwan:03:effect].


```{r}
#| label: fig-NC-corrgram
#| fig-align: center
#| fig-asp: 1
#| out-width: "90%"
#| fig-cap: "Corrgram of the `NeuroCog` data. The upper and lower triangles use two different ways of encoding the value of the correlation for each pair of variables."
NeuroCog |>
  select(-Dx, -SocialCog) |>
  corrgram(order = TRUE,
           diag.panel = panel.density,
           upper.panel = panel.pie)
```

In this plot you can see that adjacent variables are more highly correlated than those
more widely separated.  The diagonal panels show that most variables are reasonably
symmetric in their distributions. `Age`, not included in this analysis is negatively
correlated with the others: older participants tend to do less well on these tests.

### Scatterplot matrix {.unnumbered}

A scatterplot matrix gives a more detailed overview of all pairwise relations.
The plot below suppresses the data points and summarizes the relation using
data ellipses and regression lines.
The model syntax, `~ Speed + ... |Dx`, treats `Dx` as a conditioning variable
(similar to the use of the `color` aestheic in `ggplot2`)
giving a separate data ellipse and regression line for each group.
(The legend is suppressed here. The groups are: 
`r colorize("Schizophrenic", "red")`,
`r colorize("SchizoAffective", "green")`,
`r colorize("Normal", "blue")`.)

<!-- 
#| fig-cap: !expr paste("Scatterplot matrix of the `NeuroCog` data. Points are suppressed here, focusing on the data ellipses and regression lines. Colors for the groups:", colorize("Schizophrenic", "red"), colorize("SchizoAffective", "green"), colorize("Control", "blue"))

#| fig-cap: "Scatterplot matrix of the `NeuroCog` data. Points are suppressed here, focusing on the data ellipses and regression lines. Colors for the groups: Schizophrenic (red),  SchizoAffective (green), Control (blue)."

Add commas -- didn't work
#| fig-cap: !expr paste0("Scatterplot matrix of the `NeuroCog` data. Points are suppressed here, focusing on the data ellipses and regression lines. Colors for the groups: ", colorize("Schizophrenic", "red"), ", ", colorize("SchizoAffective", "green"), ", ", colorize("Control", "blue"))

-->

```{r}
#| label: fig-NC-scatmat
#| fig-align: center
#| fig-asp: 1
#| out-width: "100%"
#| fig-cap: !expr paste("Scatterplot matrix of the `NeuroCog` data. Points are suppressed here, focusing on the data ellipses and regression lines. Colors for the groups:", colorize("Schizophrenic", "red"), colorize("SchizoAffective", "green"), colorize("Control", "blue"))
col <- c("red", "darkgreen", "blue")
scatterplotMatrix(~ Speed + Attention + Memory + Verbal + Visual + ProbSolv | Dx,
                  data=NeuroCog,
                  plot.points = FALSE,
                  smooth = FALSE,
                  legend = FALSE,
                  col = col,
                  ellipse=list(levels=0.68))
```

In @fig-NC-scatmat, you can see that the regression lines have similar slopes and similar data ellipses
for the groups, though with a few exceptions. You can also see that the `r colorize("Normal", "blue")` 
group has higher means on all the variables and that the `r colorize("Schizophrenic", "red")` and 
`r colorize("SchizoAffective", "green")`  don't seem to differ very much.

### Biplot {.unnumbered}

While still in exploratory mode, we can gain greater insight with our trusty multivariate juicer,
the biplot (@sec-biplot). A PCA of the cognitive variables shows that nearly 80% of the total
variance is accounted for by two dimensions, of which the first dimension accounts for
69%. 

```{r}
neuro.pca <- NeuroCog |>
  select(Speed:Visual) |> 
  prcomp(scale. = TRUE)
summary(neuro.pca)
```

The biplot for this analysis (@fig-neuro-biplot) provides a nice summary of what was
seen in the scatterplot matrix (@fig-NC-scatmat): A large difference between the
`r colorize("Control", "blue")` group and the others, which don't differ much from each
other. 

```{r}
#| label: fig-neuro-biplot
#| out-width: "80%"
#| fig-cap: "Biplot of the NeuroCog data. The three groups differ are ordered along the first PCA dimension. Variable vectors show the correlations of the response variables with the two components."
ggbiplot(neuro.pca, 
         obs.scale = 1, var.scale = 1,
         groups = NeuroCog$Dx, 
         varname.size = 5,
         var.factor = 1.5,
         ellipse = TRUE) +
  scale_color_discrete(name = 'groups') +
  theme_minimal() +
  theme(legend.direction = 'horizontal', 
        legend.position = 'top')
```

The variable vectors in @fig-neuro-biplot reflect the correlations of the response variables
with each other and with the principal components, `PC1` and `PC2`. All of the
variables are positively correlated in this 2D view, `Memory` and `Speed` most highly so.

## Fitting the MLM

We proceed to fit the one-way MANOVA model to answer the main research question
of how well these variables distinguish among the groups.

```{r NC-lm}
NC.mlm <- lm(cbind(Speed, Attention, Memory, Verbal, Visual, ProbSolv) ~ Dx,
             data=NeuroCog)
Anova(NC.mlm)
```

The first research question is captured by the contrasts for the `Dx` factor shown above.
We can test these with `car::linearHypothesis()`.
The contrast `Dx1` for control vs. the diagnosed groups is highly significant, 

```{r NC-linhyp}
# control vs. patients
print(linearHypothesis(NC.mlm, "Dx1"), SSP=FALSE)
```

However, the second
contrast, `Dx2`, comparing the schizophrenic and schizoaffective group, shows no difference in the collection of response means.

```{r NC-linhyp2}
# Schizo vs SchizAff
print(linearHypothesis(NC.mlm, "Dx2"), SSP=FALSE)
```

As a quick check on the model, a $\chi^2$ QQ plot (@fig-NC-cqplot) reveals no problems with multivariate normality of residuals
nor potentially harmful residuals.

```{r}
#| label: fig-NC-cqplot
#| fig-width: 7
#| fig-height: 7
#| out-width: "70%"
#| fig-cap: "Chisquare QQ plot of the MANOVA model"
cqplot(NC.mlm, id.n = 3)
```

### HE plot
So the question becomes: how to understand these results.  
`heplot()` shows the visualization of the multivariate model in the space
of two response variables (the first two by default).
The result (@fig-NC-HEplot) tells a very simple story:
The control group performs higher on higher measures than the diagnosed groups, which do
not differ between themselves. 

(Because `heplot()` gets the levels of factors from the model object, in order to abbreviate the group labels in the plot, you need to `update()` the MLM model after the labels are reassigned.)

```{r}
# abbreviate levels for plots
NeuroCog$Dx <- factor(NeuroCog$Dx, 
                      labels = c("Schiz", "SchAff", "Contr"))
NC.mlm <- update(NC.mlm)
```
 
Then, feed the model to `heplot()` for a plot of the first two response variables, `Speed` and `Attention`.
```{r echo = -1}
#| label: fig-NC-HEplot
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
#| out-width: "70%"
#| fig-cap: "HE plot of Speed and Attention in the MLM for the `NeuroCog` data. The labeled points show the means of the groups on the two variables. The blue $\\mathbf{H}$ ellipse for groups indicates the strong positive correlation of the group means."
#| fig-show: hold
op <- par(mar=c(5,4,1,1)+.1)
heplot(NC.mlm, 
       fill=TRUE, fill.alpha=0.1,
       cex.lab=1.5, cex=1.35)
```

This pattern, of the control group higher than the others, is consistent across all of the response variables,
as we see from a plot of `pairs(NC.mlm)` in @fig-NC-HE-pairs.

```{r}
#| label: fig-NC-HE-pairs
#| fig-height: 9
#| fig-width: 9
#| fig-align: center
#| out-width: "100%"
#| fig-cap: "HE plot matrix of the MLM for `NeuroCog` data."
#| fig-show: hold
pairs(NC.mlm, 
      fill=TRUE, fill.alpha=0.1,
      var.cex=2)
```

The group means are nearly perfectly correlated in all panels.
This signals that we are likely to see a very simple representation of the data in canonical space.
Note that this differs from the biplot view (@fig-neuro-biplot) in that biplot ignores the
`Dx` factor in the PCA, even though we represent it in the plot.


### Canonical space

We can gain further insight, and a simplified plot showing all the response variables by projecting the MANOVA into the canonical space, which is entirely 2-dimensional (because $df_h=2$).
However, the output from `candisc()` shows that 98.5% of the mean differences among groups
can be accounted for in just one canonical dimension.
```{r NC-can}
NC.can <- candisc(NC.mlm)
NC.can
```

@fig-NC-candisc is the result of the `plot()` method for class `"candisc"` objects,
that is, the result of calling `plot(NC.can, ...)`.
It plots the two canonical scores, $\mathbf{Z}_{n \times 2}$ for the subjects,
together with data ellipses for each of the three groups.

```{r echo=-1}
#| label: fig-NC-candisc
#| fig-align: center
#| out-width: "100%"
#| fig-cap: "Canonical discriminant plot for the `NeuroCog` data MANOVA. Scores on the two canonical dimensions are plotted, together with 68% data ellipses for each group."
#| fig-show: hold
op <- par(mar=c(5,4,1,1)+.1)
pos <- c(4, 1, 4, 4, 1, 3)
col <- c("red", "darkgreen", "blue")
plot(NC.can, 
     ellipse=TRUE, 
     rev.axes=c(TRUE,FALSE), 
     pch=c(7,9,10),
     var.cex=1.2, cex.lab=1.5, var.lwd=2,  scale=4.5, 
     col=col,
	   var.col="black", var.pos=pos,
     prefix="Canonical dimension ")
```

The interpretation of @fig-NC-candisc is again fairly straightforward.
As noted earlier (@sec-candisc), the projections of the variable vectors in this plot on the coordinate axes are proportional to the correlations of the responses with the canonical scores.
From this, we see that the normal group differs from the two patient groups, having higher scores on all the neurocognitive variables, most of which are highyl correlated. The problem solving measure is slightly different, and this, compared to the cluster of memory, verbal and attention, is what distinguishes the schizophrenic group from the schizoaffectives.

The separation of the groups is essentially one-dimensional, with the control group higher on all measures. Moreover, the variables processing speed and visual memory are the purest measures of this dimension, but all variables contribute positively.
The second canonical dimension accounts for only 1.5% of group mean differences and is non-significant
(by a likelihood ratio test).  Yet, if we were to interpret it, we would note that the schizophrenia group is slightly higher on this dimension, scoring better in problem solving and slightly worse on working memory, attention, and verbal learning tasks.

**Summary**

This analysis gives a very simple description of the data, in relation to the research questions posed earlier:

* On the basis of these neurocognitive tests, the schizophrenic and schizoaffective groups do not differ significantly overall, but these groups differ greatly from the normal controls.

* All cognitive domains distinguish the groups in the same direction, with the greatest differences shown for the variables most closely aligned with the horizontal axis in @fig-NC-candisc. This means that these neuro-psychological
measures can be considered to tap a single dimension of differences among the three diagnostic groups.

## Social cognitive measures {#sec-social-cog}

The social cognitive measures in the @Hartman:2016 study
were designed to tap various aspects of the perception and cognitive
processing of emotions of others.  Emotion perception was assessed using a Managing Emotions
score from the MCCB.  A "theory of mind" (`ToM`) score assessed ability to read the emotions of
others from photographs of the eye region of male and female faces. Two other measures,
externalizing bias (`ExtBias`) and personalizing bias (`PersBias`) were
calculated from a scale measuring the degree to which individuals attribute internal, personal
or situational causal attributions to positive and negative social events.

The analysis of the `SocialCog` data proceeds in a similar way: first we fit the MANOVA model,
then test the overall differences among groups using `Anova()`.  We find that the overall multivariate test is again significant, 


```{r SC-mlm}
data(SocialCog, package="heplots")
SC.mlm <-  lm(cbind(MgeEmotions,ToM, ExtBias, PersBias) ~ Dx,
               data=SocialCog)
Anova(SC.mlm)
```

Testing the same
two contrasts using `linearHypothesis()` (results not shown), w
e find that the overall multivariate test is again significant, but now _both_ contrasts
are significant (Dx1: $F(4, 133)=5.21, p < 0.001$; Dx2: $F(4, 133)=2.49, p = 0.0461$),
the test for `Dx2` just barely.

```{r SC-mlm1}
#| results: hide
# control vs. patients
print(linearHypothesis(SC.mlm, "Dx1"), SSP=FALSE)
# Schizo vs. SchizAff
print(linearHypothesis(SC.mlm, "Dx2"), SSP=FALSE)
```

These results are important, because, if they are reliable and make sense
substantively, they imply that patients with
schizophrenia and schizoaffective diagnoses _can_ be distinguished by their performance
on tasks assessing social perception and cognition. This was potentially a new finding in the
literature on schizophrenia.


As we did above, it is useful to visualize the nature of these differences among groups
with HE plots for the `SC.mlm` model.
Each contrast has a corresponding $\mathbf{H}$ ellipse, which we can show in the plot
using the `hypotheses` argument. With a single degree of freedom, these degenerate
ellipses plot as lines.

```{r echo=-1}
#| label: fig-SC-HEplot
#| fig-align: center
#| out-width: "70%"
#| fig-cap: "HE plot of Speed and Attention in the MLM for the `SocialCog` data. The labeled points show the means of the groups on the two variables. The lines for Dx1 and Dx2 show the tests of the contrasts among groups."
#| fig-show: hold
op <- par(mar=c(5,4,1,1)+.1)
heplot(SC.mlm, 
       hypotheses=list("Dx1"="Dx1", "Dx2"="Dx2"),
       fill=TRUE, fill.alpha=.1,
       cex.lab=1.5, cex=1.2)
```

It can be seen that the three group
means are approximately equally spaced on the `ToM` measure, whereas for
`MgeEmotions`, the control and schizoaffective groups are quite similar, and
both are higher than the schizophrenic group.  This ordering of the three
groups was somewhat similar for the other responses, as we could see in a
`pairs(SC.mlm)` plot.


### Model checking

Normally, we would continue this analysis, and consider other HE and canonical
discriminant plots to further interpret the results, in particular the
relations of the cognitive measures to group differences, or perhaps an
analysis of the relationships between the neuro- and social-cognitive measures.
We don't pursue this here for reasons of length, but this example actually has a more important lesson to demonstrate.

Before beginning the MANOVA analyses, extensive data screening was done by the client
using SPSS, in which all the response _and_ predictor variables were checked for univariate normality
and multivariate normality (MVN) for both sets.
This traditional approach yielded a huge amount of tabular output and no
graphs, and did not indicate any major violation of assumptions.[^mvn]

[^mvn]: Actually, multivariate normality of the predictors in $\mathbf{X}$ is not required in the MLM.
This assumption applies only to the conditional values $\mathbf{Y} \;|\; \mathbf{X}$, i.e., that the
errors $\mathbf{\epsilon}_{i}' \sim \mathcal{N}_{p}(\mathbf{0},\boldsymbol{\Sigma})$
with constant covariance matrix.  Moreover, the widely used MVN test statistics, such as 
Mardia's [-@Mardia:1970:MMS] test based on multivariate skewness and kurtosis are known to be quite
sensitive to mild departures in kurtosis [@Mardia:1974] which do not threaten the validity of the multivariate tests.

A simple visual test of MVN and the possible presence of multivariate outliers is related to
the theory of the data ellipse: Under MVN, the squared Mahalanobis distances
$D^2_M (\mathbf{y}) = (\mathbf{y} - \bar{\mathbf{y}})' \, \mathbf{S}^{-1} \, (\mathbf{y} - \bar{\mathbf{y}})$
should follow a $\chi^2_p$ distribution. Thus, a quantile-quantile plot of the
ordered $D^2_M$ values vs. corresponding quantiles of the $\chi^2$ distribution
should approximate a straight line [@Cox:1968; @Healy:1968:MNP]. 
Note that this should be applied to the _residuals_
from the model -- `residuals(SC.mlm)` -- and not to the response variables directly.

`heplots::cqplot()` implements this for `"mlm"` objects
Calling this function for the model `SC.mlm` produces @fig-SC-cqplot.
It is immediately apparent that there is one extreme multivariate outlier; three other
points are identified, but the remaining observations
are nearly within the 95% confidence envelope (using a robust MVE estimate of $\mathbf{S}$).

```{r echo=-1}
#| label: fig-SC-cqplot
#| fig-align: center
#| out-width: "70%"
#| fig-cap: "Chi-square quantile-quantile plot for residuals from the model `SC.mlm`. The confidence band gives a point-wise 95% envelope, providing information about uncertainty. One extreme multivariate outlier is highlighted."
#| fig-show: hold
op <- par(mar=c(5,4,1,1)+.1)
cqplot(SC.mlm, method="mve", 
       id.n=4, 
       main="", 
       cex.lab=1.25)
```

Further checking revealed that this was a data entry error where one case (15) in the
schizophrenia group had a score of -33
recorded on the `ExtBias` measure, whose valid range was (-10, +10). In R, it is very easy to
re-fit a model to a subset of observations (rather than modifying the dataset itself) using
`update()`.  The result of the overall Anova and the test of `Dx1` were unchanged;
however, the multivariate test for the most interesting contrast `Dx2` comparing the
schizophrenia and schizoaffective groups became non-significant at the $\alpha=0.05$ level
($F(4, 133)=2.18, p = 0.0742$).

```{r SC-update, results='hide'}
SC.mlm1 <- update(SC.mlm, 
                  subset=rownames(SocialCog)!="15")

Anova(SC.mlm1)
print(linearHypothesis(SC.mlm1, "Dx1"), SSP=FALSE)
print(linearHypothesis(SC.mlm1, "Dx2"), SSP=FALSE)
```


### Canonical HE plot

This outcome creates a bit of a quandry for further analysis 
(do univariate follow-up tests? try a robust model?) and reporting (what to claim about the `Dx2` contrast?) that we don't explore here.
Rather, we proceed to attempt to interpret the MLM with the aid of canonical analysis and
a canonical HE plot.  The canonical analysis of the model `SC.mlm1` now shows that
both canonical dimensions are significant, and account for 83.9% and 16.1% of between
group mean differences respectively.

```{r SC-can1}
SC.can1 <- candisc(SC.mlm1)
SC.can1
```

```{r}
#| label: fig-SC1-hecan
#| fig-align: center
#| out-width: "70%"
#| fig-cap: "Canonical HE plot for the corrected `SocialCog` MANOVA. The variable vectors show the correlations of the responses with the canonical variables. The embedded green lines show the projections of the **H** ellipses for the contrasts `Dx1` and `Dx2` in canonical space."
op <- par(mar=c(5,4,1,1)+.1)
heplot(SC.can1, 
	fill=TRUE, fill.alpha=.1,
	hypotheses=list("Dx1"="Dx1", "Dx2"="Dx2"),
	lwd = c(1, 2, 3, 3),
	col=c("red", "blue", "darkgreen", "darkgreen"),
	var.lwd=2, 
	var.col="black", 
	label.pos=c(3,1), 
	var.cex=1.2, 
	cex=1.25, cex.lab=1.2, 
	scale=2.8,
	prefix="Canonical dimension ")
par(op)
```

The HE plot version of this canonical plot is shown in @fig-SC1-hecan. Because the
`heplot()` method for a `"candisc"` object refits the original model to the
$\mathbf{Z}$ canonical scores, it is easy to also project other linear hypotheses into
this space.  Note that in this view, both the `Dx1` and `Dx2` contrasts project outside
$\mathbf{E}$ ellipse.^[The direct application of significance tests to canonical scores probably requires some adjustment because these are computed to have the optimal between-group discrimination.].


This canonical HE plot has a very simple description:

* Dimension 1 orders the groups from control to schizoaffective to schizophrenia, while dimension 2 separates the schizoaffective group from the others;
* Externalizing bias and theory of mind contributes most to the first dimension, while personal bias and managing emotions are more aligned with the second; and,
* The relations of the two contrasts to group differences and to the response variables can be easily read from this plot.






```{r}
#cat("Packages used here:\n")
write_pkgs(file = .pkg_file)
```

<!-- ## References {.unnumbered} -->
