```{r include=FALSE}
source("R/common.R")
knitr::opts_chunk$set(fig.path = "figs/exercises/")
```

::: {.content-visible unless-format="pdf"}
{{< include latex/latex-commands.qmd >}}
:::

# Exercises {#sec-exercises}

This appendix collects exercises for each chapter of the book.

## Chapter 1: Warm-up Exercises {.unnumbered}

## Chapter 2: Introduction {.unnumbered}

## Chapter 3: Getting Started {.unnumbered}

## Chapter 4: Plots of Multivariate Data {.unnumbered}

```{r child="exercises/Ch04-exercises.qmd"}
```

## Chapter 5: Dimension Reduction {.unnumbered}

## Chapter 6: Overview of Linear models {.unnumbered}

## Chapter 7: Plots for Univariate Response Models {.unnumbered}

## Chapter 8: Topics in Linear Models {.unnumbered}

## Chapter 9: Collinearity & Ridge Regression {.unnumbered}

## Chapter 10: Hotelling's $T^2$ {.unnumbered}

## Chapter 11: Multivariate Linear Models {.unnumbered}

## Chapter 12: Visualizing Multivariate Models {.unnumbered}

## Chapter 13: Visualizing Equality of Covariance Matrices {.unnumbered}

## Chapter 14: Multivariate Influence and Robust Estimation {.unnumbered}

## Chapter 15 (Appendix): Discriminant Analysis {.unnumbered}
