```{r include=FALSE}
source("R/common.R")
```

# Colophon {.unnumbered}

This book was produced using [`r R.version.string`](https://cran.r-project.org/). Fundamental to this was the framework for reproducible documents provided by Yihui Xie's [knitr package](https://cran.r-project.org/web/packages/knitr/index.html).

The book was entirely written within the [RStudio](https://posit.co/download/rstudio-desktop) IDE environment, which made it easy to develop the analyses
and graphs, and then write about them, inserting code chunks, figures, cross-references, and so forth.
[Quarto](https://quarto.org/) was used to compile and render the book in HTML and PDF formats, with the helpers
of 

* [knitr](https://cran.r-project.org/web/packages/knitr/index.html) to run code chunks and incorporate into document;
* [pandoc](https://pandoc.org/) to convert among various outputs;
* [TinyTeX](https://yihui.org/tinytex/) to provide a coherent set of LaTeX tools for PDF output


```{r}
#| label: quarto-info
#| echo: false

tinytex_ver <- tryCatch({
  v <- system2("tlmgr", "--version", stdout = TRUE, stderr = FALSE)[1]
  m <- regmatches(v, regexpr("[0-9]{4}", v))
  if (length(m)) m else "?"
}, error = function(e) "?")

data.frame(
  Tool    = c("Quarto", "Pandoc", "R", "knitr", "rmarkdown", "TinyTeX"),
  Version = c(
    system2("quarto", "--version", stdout = TRUE)[1],
    as.character(rmarkdown::pandoc_version()),
    paste(R.version$major, R.version$minor, sep = "."),
    as.character(packageVersion("knitr")),
    as.character(packageVersion("rmarkdown")),
    tinytex_ver
  )
) |> knitr::kable(row.names = FALSE, align = c("l", "r"))
```

## Package versions

The principal R package versions used in examples and illustrations are listed below. 
These were captured via `sessioninfo:::package_info()`
from all `library()` commands in the text, and scripts which also updated
the references to packages.

At the time of writing, most of these were current on [CRAN](http://cran.us.r-project.org/)
repositories, but some development versions are listed with their download repo, or indicated as "local" in the `source` column. 

```{r}
#| label: package-info
#| echo: false
#| results: asis
library(dplyr)

make_source_link <- function(pkg, source) {
  if (grepl("^CRAN", source)) {
    url <- paste0("https://cran.r-project.org/package=", pkg)
    paste0("[CRAN](", url, ")")
  } else if (grepl("r-universe", source, ignore.case = TRUE)) {
    base <- sub("/$", "", source)  # strip trailing slash if present
    paste0("[R-universe](", base, "/", pkg, ")")
  } else if (grepl("^Github", source)) {
    repo <- sub("^Github \\(([^)]+)\\).*", "\\1", source)
    paste0("[", repo, "](https://github.com/", repo, ")")
  } else {
    source
  }
}

packages <- read_pkgs()
pkg_info <- sessioninfo:::package_info(packages, dependencies = FALSE)
pkg_info$source <- sub(" \\(R.*\\)", "", pkg_info$source)
pkg_info$source <- sub("@[0-9a-f]{7,}", "", pkg_info$source)
pkg_info |>
  select(package, ondiskversion, source) |>
  rename(version = ondiskversion) |>
  mutate(source = mapply(make_source_link, package, source)) |>
  knitr::kable(format = "pipe", row.names = FALSE, align = c("l", "r", "l"))
```

