```{r include=FALSE}
source("R/common.R")
```

# Colophon {.unnumbered}

This book was produced using [`r R.version.string`](https://cran.r-project.org/). Fundamental to this was the framework for reproducible documents provided by Yihui Xie's [knitr package](https://cran.r-project.org/web/packages/knitr/index.html).

[Quarto](https://quarto.org/) was used to compile and render the book in HTML and PDF formats.
[** Don't really need all this**]

```{r}
#| label: quarto-info
#| echo: false
#| comment: ""

# this from: https://raw.githubusercontent.com/aml4td/website/main/index.qmd
quarto_info <- function(){
  file_out <- tempfile("temp-quarto.txt")
  system2(command = "quarto", args = "check", stderr = file_out)
  res <- readLines(file_out)
  res <- purrr::map_chr(res, cli::ansi_strip)
  rms <- c("(|)", "(/)", "(\\)", "(/)", "(-)", "/Users", "Path:", "Install with")
  for (pat in rms) {
    res <- res[!grepl(pat, res, fixed = TRUE)]
  }
  res <- res[res != ""]
  invisible(res)
}

quarto_res <- quarto_info()
req_quarto_version <- "1.4.510"  # in future, use numeric_version()

# if (!any(grepl(req_quarto_version, quarto_res))) {
#   cli::cli_abort("Version {req_quarto_version} of Quarto is required. 
#                   See {.url https://quarto.org/docs/download/}")
# }

cat(quarto_res, sep = "\n")
```

## Package versions

The principal R package versions used in examples and illustrations are listed below. 
These were captured via `sessioninfo:::package_info()`
from all `library()` commands in the text, and scripts which also updated
the references to packages.

At the time of writing, most of these were current on [CRAN](http://cran.us.r-project.org/)
repositories but some development versions are indicated as "local" in the `source` column. 

```{r}
#| label: package-info
#| echo: false
#| results: asis
library(dplyr)

packages <- read_pkgs()
pkg_info <- sessioninfo:::package_info(packages, dependencies = FALSE) 
# clean up unwanted
pkg_info$source <- sub(" \\(R.*\\)", "", pkg_info$source)
#pkg_info <- pkg_info[,-2]
pkg_info |>
  select(package, ondiskversion, date, source) |>
  rename(version = ondiskversion) |>
  knitr::kable(row.names = FALSE)

#print(pkg_info)
```

