```{r include=FALSE}
source("R/common.R")
```

**Packages**

In this chapter we use the following packages. Load them now
```{r}
library(car)
library(heplots)
library(candisc)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Visualizing Tests for Equality of Covariance Matrices

This chapter concerns the extension of tests of homogeneity of variance from the classical univariate
ANOVA setting to the analogous multivariate (MANOVA) setting. Such tests are a routine but important aspect of data analysis, as particular violations can drastically impact model estimates [@Lix:1996]. 


We provide some answers to the following questions:

- **Visualization**: How can we visualize differences among group variances and covariance matrices,
  perhaps in a way that is analogous to what is done to visualize differences among group means?
  As will be illustrated, differences among covariance matrices can be comprised of spread in overall size ("scatter") and shape ("orientation"). These can be seen in data space with data ellipses, particularly if the data is centered by shifting all groups to the grand mean,
  
- **Low-D views**:  When there are more than a few response variables, what low-dimensional
  views can show the most interesting properties related to the equality of covariance matrices?
  Projecting the data into the space of the principal components serves well again here. Surprisingly, we will see that the small dimensions contain useful information about differences among the group covariance matrices.
  
- 



