```{r include=FALSE}
source("R/common.R")
```

**Packages**

In this chapter we use the following packages. Load them now
```{r}
library(car)
library(heplots)
library(candisc)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Visualizing Tests for Equality of Covariance Matrices

> _To make the preliminary test on variances is rather like putting to sea in a rowing boat
> to find out whether conditions are sufficiently calm for an ocean liner to leave port._
> --- G. E. P. Box [-@Box:1953]


This chapter concerns the extension of tests of homogeneity of variance from the classical univariate
ANOVA setting to the analogous multivariate (MANOVA) setting. Such tests are a routine but important aspect of data analysis, as particular violations can drastically impact model estimates [@Lix:1996]. 


We provide some answers to the following questions:

- **Visualization**: How can we visualize differences among group variances and covariance matrices,
  perhaps in a way that is analogous to what is done to visualize differences among group means?
  As will be illustrated, differences among covariance matrices can be comprised of spread in overall size ("scatter") and shape ("orientation"). These can be seen in data space with data ellipses, particularly if the data is centered by shifting all groups to the grand mean,
  
- **Low-D views**:  When there are more than a few response variables, what low-dimensional
  views can show the most interesting properties related to the equality of covariance matrices?
  Projecting the data into the space of the principal components serves well again here. Surprisingly, we will see that the small dimensions contain useful information about differences among the group covariance matrices.
  
- **Other statistics**: Box's $M$-test is most widely used. Are there other worthwhile test statistics? We will see that graphics methods suggest alternatives.

The following subsections provide a capsule summary of the issues in this topic.
Most of the discussion is couched in terms of a one-way design for simplicity,
but the same ideas can apply to two-way (and higher) designs, where a "group" factor
is defined as the product combination (interaction) of two or more factor
variables. When there are also numeric covariates, this topic can also be extended
to the multivariate analysis of covaraiance (MANCOVA) setting. This can be 
accomplished by applying these techniques to the residuals from predictions by 
the covariates alone.

## Homogeneity of Variance in Univariate ANOVA

In classical (Gaussian) univariate ANOVA models, the main interest is typically on
tests of mean differences in a response $y$ according to one or more factors.
The validity of the typical $F$ test, however, relies on the assumption
of _homogeneity of variance_: 
all groups have the same (or similar) variance,
$$
\sigma_1^2 = \sigma_2^2 = \cdots = \sigma_g^2 \; .
$$

It turns out that the $F$ test for differences in means
is relatively robust to violation of this assumption [@Harwell:1992],
as long as the group sizes are roughly equal.[^1]

[^1]: If group sizes are greatly unequal __and__ homogeneity of variance is violated, then the $F$ statistic is too liberal ($p$ values too large) when large sample variances are associated with small group sizes.
Conversely, the $F$ statistic is too conservative if large variances are associated with large group sizes.

A variety of classical test statistics for homogeneity of variance are available, including
Hartley's $F_{max}$ [@Hartley:1950], Cochran's _C_ [@Cochran:1941],and Bartlett's test [@Bartlett:1937],
but these have been found to have terrible statistical properties [@Rogan:1977], which prompted Box's famous quote.

Levene [-@Levene:1960] introduced a different form of test, based on the simple idea that when 
variances are equal across groups, 
the average absolute values of differences between the observations and group means
will also be equal, i.e., substituting an $L_1$ norm for the $L_2$ norm of variance. 
In a one-way design, this is equivalent to a test of group differences in the
means of the auxilliary variable $z_{ij} = | y_{ij} - \bar{y}_i |$.

More robust versions of this test were proposed by @BrownForsythe:1974. These tests
substitute the group mean by either the group median or a trimmed mean in the ANOVA of the absolute deviations, and
should be almost always preferred to Levene's version.
See @Conover-etal:1981 for an early review and @Gastwirth-etal:2009 for a general discussion of these tests. 
In what follows, we refer to this class of tests as "Levene-type" tests and suggest a 
multivariate extension described below (@sec-mlevene).

## Homogeneity of variance in MANOVA

In the MANOVA context, the main emphasis, of course, is on differences among mean vectors,
testing
$$
H_0 : \mathbf{\mu}_1 = \mathbf{\mu}_2 = \cdots = \mathbf{\mu}_g \; .
$$
However, the standard test statistics (Wilks' Lambda, Hotelling-Lawley trace, Pillai-Bartlett trace, Roy's maximum root) rely upon
the analogous assumption that
the within-group covariance matrices for all groups are equal,
$$
\mathbf{\Sigma}_1 = \mathbf{\Sigma}_2 = \cdots = \mathbf{\Sigma}_g \; .
$$

**Insert** pairs covEllipses for penguins data

To preview the main example, @fig-peng-covEllipse0 shows data ellipses for the
main size variables in the `palmerpenguins::penguins` data.

```{r}
#| label: fig-peng-covEllipse0
#| fig-align: center
#| out-width: "100%"
#| code-fold: show
#| echo: FALSE
#| fig-cap: "Data ellipses for the penguins data."

load(here::here("data", "peng.RData"))

covEllipses(peng[3:6], peng$species, 
            variables=1:4,
            pooled = FALSE,
            fill=c(rep(FALSE,3), TRUE), 
            fill.alpha=.1)
```

They covariance ellipses look pretty similar in size, shape and orientation.
But what does Box's M test (described below) say? As you can see,
it concludes strongly against the null hypothesis.

```{r}
boxM(cbind(bill_length, bill_depth, flipper_length, body_mass) ~ species, data=peng)
```


## Assessing heterogeneity of covariance matrices: Box's M test

@Box:1949 proposed the following likelihood-ratio test (LRT) statistic for testing the
hypothesis of equal covariance matrices,
$$
M = (N -g) \ln \;|\; \mathbf{S}_p \;|\; - \sum_{i=1}^g (n_i -1) \ln \;|\; \mathbf{S}_i \;|\; \; ,
$$ {eq-boxm}

where $N = \sum n_i$ is the total sample size and
$\mathbf{S}_p = (N-g)^{-1} \sum_{i=1}^g (n_i - 1) \mathbf{S}_i$ 
is the pooled covariance matrix.
$M$ can thus be thought of as a ratio of the determinant of the pooled $\mathbf{S}_p$
to the geometric mean of the determinants of the separate $\mathbf{S}_i$.

In practice, there are various transformations of the value of $M$ to yield a
test statistic with an approximately known distribution [@Timm:75].
Roughly speaking, when each $n_i > 20$, a $\chi^2$ approximation is often used;
otherwise an $F$ approximation is known to be more accurate.

Asymptotically, $-2 \ln (M)$ has a $\chi^2$ distribution. The $\chi^2$ approximation
due to Box [-@Box:1949; -@Box:1950] is that
$$
X^2 = -2 (1-c_1) \ln (M) \quad \sim \quad \chi^2_{df}
$$
with $df = (g-1) p (p+1)/2$ degrees of freedom, and a bias correction constant:
$$
c_1 = \left( 
\sum_i \frac{1}{n_i -1}
- \frac{1}{N-g}
\right)
\frac{2p^2 +3p -1}{6 (p+1)(g-1)} \; .
$$

In this form, Bartlett's test for equality of variances in the univariate case
is the special case of Box's M when there is only one response variable,
so Bartlett's test is sometimes used as univariate follow-up to determine
which response variables show heterogeneity of variance.

Yet, like its univariate counterpart, Box's test is well-known to be highly sensitive to violation
of (multivariate) normality and the presence of outliers. For example, @TikuBalakrishnan:1984
concluded from simulation studies that the normal-theory LRT provides poor control of Type I
error under even modest departures from normality. @OBrien:1992 proposed some robust alternatives,
and showed that Box's normal theory approximation suffered both in controlling the null size
of the test and in power. @ZhangBoos:1992:BCV also carried out simulation studies with similar
conclusions and used bootstrap methods to obtain corrected critical values.

## Visualizing heterogeneity

The goal of this chapter is to use the above background as a platform for discussing approaches to visualizing and testing the heterogeneity of covariance matrices in multivariate designs.  While researchers often rely on a single number to determine if their data have met a particular threshold, such compression will often obscure interesting information, particularly when a test concludes that differences exist, and one is left to wonder ``why?''.
It is within this context where, again, visualizations often reign supreme. 
In fact, we find it somewhat surprising that this issue has not been addressed before graphically in any systematic way. **TODO: cut this down**

In what follows, we propose three visualization-based approaches to questions of heterogeneity of covariance in MANOVA
designs:

(a) direct visualization of the information in the $\mathbf{S}_i$ and $\mathbf{S}_p$ using _data ellipsoids_ to show size and shape as minimal schematic summaries;

(b) a simple dotplot of the components of Box's M test: the log determinants of the 
$\mathbf{S}_i$ together with that of the pooled $\mathbf{S}_p$. Extensions of these simple plots raise the question of whether measures of heterogeneity other than that captured in Box's test might also be useful; and,

(c) the connection between Levene-type tests and an ANOVA (of centered absolute differences) suggests a parallel with a multivariate extension of Levene-type tests and a MANOVA.  We explore this with a version of
Hypothesis-Error (HE) plots we have found useful for visualizing mean differences in MANOVA designs.

```{r echo=FALSE}
cat("Writing packages to ", .pkg_file, "\n")
write_pkgs(file = .pkg_file)
```

## References
