```{r include=FALSE}
source("R/common.R")
clean_pkgs()
```


# Preface {.unnumbered}

This is an early draft of material that may or may not appear in the preface.

```{r child="child/one-two-many.qmd"}
```


```{r child="child/flatland.qmd"}
```

```{r child="child/eureka.qmd"}
```

## What I assume
I assume the reader to have at least a basic familiarity with R.
While R fundamentals are outside the scope of the current paper, I believe that this language provides a rich set of resources, far beyond that offered by other statistical software packages, and is well worth learning.

For those not familiar with R, I recommend @Matloff-2011, @Wickham2014, and @Cotton-2013 for introductions to programming in the language, and @FoxWeisberg:2018 and @Teetor2011 for learning about how to conduct basic statistical analyses.



## Conventions used in this book

The following typographic conventions are used in this book:

* _italic_ : indicates terms to be _emphasized_ or defined in the text, ...

* **bold** : is used for names of R packages (well, not so far)

* `fixed-width` : is used in program listings as well as in text to refer to variable and function names, R statement elements and keywords.

* _`fixed-width italic`_ : isn't used yet, but probably should be.

For R functions in packages, we use the notation `package::function()`, for example: `car::Anova()` to identify where those functions are defined



## References {.unnumbered}
