```{r include=FALSE}
source("R/common.R")
```

# Plots for univariate response models {#sec-linear-models-plots}

For a univariate linear model fit using `lm()`, `glm()` and similar functions, the standard `plot()`
method gives basic versions of plots of residuals and other calculated quantities for assessing
possible violations of the model assumptions.
Some of these can be considerably enhanced using other packages.

Beyond this, 

* tables of model coefficients, standard errors and test statistics can often be usefully 
supplemented or even replaced by suitable plots providing essentially the same information.

* when there are two or more predictors, you can more easily understand their separate impact
on the response by plotting the _marginal_ effects of one or more focal variables, averaging
over other variables not shown in a given plot.

* when there are highly correlated predictors, some specialized plots are useful to
understand the nature of _multicolinearity_.

## The "regression quartet"

## Other Diagnostic plots

## Coefficient plots

## Added-variable plots 

## Marginal plots

## Visualizing multicolinearity

```{r echo=FALSE}
cat("Writing packages to ", .pkg_file, "\n")
write_pkgs(file = .pkg_file)
```

## References {.unnumbered}
