### A real example {#sec-davis}

In the mid 1980s, a consulting client had a strange problem.
She was conducting a study of the relation between body image and weight preoccupation in
exercising and non-exercising people [@Davis:1990]. As part of the design, the researcher
wanted to know if self-reported weight could be taken as a reliable indicator of
true weight measured on a scale. It was expected that the correlations between reported
and measured weight should be close to 1.0, and the slope of the regression lines
for men and women should also be close to 1.0. The data set is `car::Davis`.

She was therefore very surprise to see the following numerical results: For men,
the correlation was nearly perfect, but not so for women.

```{r davis-cor}
data(Davis, package="carData")
Davis <- Davis |>
  drop_na()          # drop missing cases
Davis |>
  group_by(sex) |>
  select(sex, weight, repwt) |>
  summarise(r = cor(weight, repwt))
```

Similarly, the regression lines showed the expected slope for men, but that for women was only 0.26.

```{r davis-lm}
Davis |>
  nest(data = -sex) |>
  mutate(model = map(data, ~ lm(repwt ~ weight, data = .)),
         tidied = map(model, tidy)) |>
  unnest(tidied) |>
  filter(term == "weight") |>
  select(sex, term, estimate, std.error)

```

What could be wrong here?, the client asked. The consultant replied with the obvious question:

> _Did you plot your data?_

The answer turned out to be one discrepant point, a female, whose measured weight was 166 kg
(366 lbs!). This single point exerted so much influence that it pulled the fitted regression line
down to a slope of only 0.26.

```{r}
#| label: fig-ch02-davis-reg1
#| fig-cap: "Regression for Davis' data on reported weight and measures weight for men and women.
#|    Separate regression lines, predicting reported weight from measured weight are shown for males and females.
#|    One highly unusual point is highlighted."
Davis |>
  ggplot(aes(x = weight, y = repwt, 
             color = sex, shape=sex)) +
  geom_point(size = ifelse(Davis$weight==166, 6, 2)) +
  labs(y = "Measured weight (kg)", 
       x = "Reported weight (kg)") +
  geom_smooth(method = "lm", formula = y~x, se = FALSE) +
  theme(legend.position = c(.8, .8))
```

In this example, it was arguable that $x$ and $y$ axes should be reversed, 
to determine how well measured weight can be predicted from reported weight.
In `ggplot` this can easily be done by reversing the `x` and `y` aesthetics.

```{r}
#| label: fig-ch02-davis-reg2
#| fig-cap: "Regression for Davis' data on reported weight and measures weight for men and women.
#|    Separate regression lines, predicting measured weight from re[ported] weight are shown for males and females.
#|    The highly unusual point no longer has an effect on the fitted lines."
Davis |>
  ggplot(aes(y = weight, x = repwt, color = sex, shape=sex)) +
  geom_point(size = ifelse(Davis$weight==166, 6, 2)) +
  labs(y = "Measured weight (kg)", x = "Reported weight (kg)") +
  geom_smooth(method = "lm", formula = y~x, se = FALSE) +
  theme(legend.position = c(.8, .8))
```

In @fig-ch02-davis-reg2, this discrepant observation again stands out like a sore thumb, but
it makes very little difference in the fitted line for females.
The reason is that this point is well within the range of the $x$ variable (`repwt`).
To impact the slope of the regression line, an observation must be unusual in_both_ $x$ and $y$.
We take up the topic of how to detect influential observations and what to do about them in 
Chapter XX.

The value of such plots is not only that they can reveal possible problems with an analysis,
but also help identify their reasons and suggest corrective action.
What went wrong here? Examination of the original data showed that this person switched
the values, recording her reported weight in the box for measured weight and vice versa.
