
<!-- **Packages** -->
<!-- In this chapter we use the following packages. Load them now -->
<!-- ```{r} -->
<!-- library(car) -->
<!-- library(heplots) -->
<!-- library(Hotelling) -->
<!-- library(ggplot2) -->
<!-- library(dplyr) -->
<!-- library(tidyr) -->
<!-- ``` -->

## Neuro- and Social-cognitive measures in psychiatric groups

A Ph.D. dissertation by Laura Hartman [-@Hartman:2016] at York University was designed to evaluate whether and
how clinical patients diagnosed (on the DSM-IV) as schizophrenic or with schizoaffective disorder could be
distinguished from each other and from a normal, control sample on collections of standardized
tests in the following domains:

* **Neuro-cognitive**: processing speed, attention, verbal learning, visual learning and problem solving;

* **Social-cognitive**: managing emotions, theory of mind, externalizing, personalizing bias.

The study is an important contribution to clinical research because the two diagnostic categories
are subtly different and their symptoms often overlap. Yet, they’re very different and often require different treatments. A key difference between schizoaffective disorder and schizophrenia is the prominence of mood disorder involving bipolar, manic and depressive moods.
With schizoaffective disorder, mood disorders are front and center. With schizophrenia, that is not a dominant part of the disorder, but psychotic ideation (hearing voices, seeing imaginary people) is.

### Research questions

This example is concerned with the following substantitive questions:

* To what extent can patients diagnosed as schizophrenic or with schizoaffective disorder  be distinguished from each other and from a normal control sample using a well-validated, comprehensive neurocognitive battery specifically designed for individuals with psychosis  [@Heinrichs-etal:2015] ?

* If the groups differ, do any of the cognitive domains show particularly larger or smaller differences among these groups? 

* Do the neurocognitive discriminate among the in the same or different ways?
If different, how many separate aspects or dimensions are distinguished?

It could aid diagnosis and treatment if these similar mental disorders could be distinguished by tests in the cognitive domain.


### Data

The clinical sample comprised 116 male and female patients who met the following criteria: 1) a diagnosis of schizophrenia ($n$ = 70) or schizoaffective disorder ($n$ = 46) confirmed by the Structured Clinical Interview for DSM-IV-TR Axis I Disorders; 2) were outpatients; 3) a history free of developmental or learning disability; 4) age 18-65; 5) a history free of neurological or endocrine disorder; and 6) no concurrent diagnosis of substance use disorder.
Non-psychiatric control participants ($n$ = 146) were screened for medical and psychiatric illness and history of substance abuse and were recruited from three outpatient clinics.

```{r NC-data}
data(NeuroCog, package="heplots")
glimpse(NeuroCog)
```

The diagnostic classification variable is called `Dx` in the data set. To facilitate answering questions regarding group differences, the following contrasts were applied:
the first column compares the control group to the average of the diagnosed groups, the second
compares the schizophrenia group against the schizoaffective group.

```{r NC-contrasts}
contrasts(NeuroCog$Dx)
```


In this analysis, we ignore the `SocialCog` variable. The primary focus is on
the variables `Attention : ProbSolv`.

### A first look

As always, plot the data first! We want an overview of the distributions of the variables to see
the centers, spread,  shape and possible outliers for each group on each variable.

The plot below combines the use of boxplots and violin plots to give an informative display.
As we saw earlier (Chapter XXX), doing this with `ggplot2 requires reshaping the data to
long format.

```{r NC-long}
# Reshape from wide to long
NC_long <- NeuroCog |>
  dplyr::select(-SocialCog, -Age, -Sex) |>
  tidyr::gather(key = response, value = "value", Speed:ProbSolv)
# view 3 observations per group and measure
NC_long |>
  group_by(Dx) |>
  sample_n(3) |> ungroup()
```

In the plot, we take care to adjust the transparency (`alpha`) values for the points,
violin plots and boxplots so that all can be seen. Options for `geom_boxplot()`
are used to give these greater visual prominence.
```{r}
#| label: fig-NC-boxplot
#| fig-align: center
#| out-width: "120%"
#| fig-cap: "Boxplots and violin plots of the `NeuroCog` data."
ggplot(NC_long, aes(x=Dx, y=value, fill=Dx)) +
  geom_jitter(shape=16, alpha=0.8, size=1, width=0.2) +
  geom_violin(alpha = 0.1) +
  geom_boxplot(width=0.5, alpha=0.4, 
               outlier.alpha=1, outlier.size = 3, outlier.color = "red") +
  scale_x_discrete(labels = c("Schizo", "SchizAff", "Control")) +
  facet_wrap(~response, scales = "free_y", as.table = FALSE) +
  theme_bw() +
  theme(legend.position="bottom",
        axis.title = element_text(size = rel(1.2)),
        axis.text  = element_text(face = "bold"),
        strip.text = element_text(size = rel(1.2)))
```

We can see that the control participants score higher on all measures, but there is no consistent
pattern of medians for the two patient groups. But these univariate summaries do not inform
about the relations among variables.

### Bivariate views

A corrgram [@Friendly:02:corrgram] provides a useful reconnaisance plot of the bivariate correlations in the data set.It suppresses details, and allows focus on the overall pattern.
The `corrgram::corrgram()` function has the ability to enhance perception by
permuting the variables in the order of their variable vectors in a biplot.

The plot below includes all variables except for `Dx` group.
There are a number of `panel.*` functions for choosing how the correlation for each pair is
rendered.

```{r}
#| label: fig-NC-corrgram
#| fig-align: center
#| fig-asp: 1
#| out-width: "90%"
#| fig-cap: "Scatterplot matrix of the `NeuroCog` data. Points are suppressed here, focusing on the data ellipses and regression lines"
NeuroCog |>
  select(-Dx) |>
  corrgram(order = TRUE,
           diag.panel = panel.density,
           upper.panel = panel.pie)
```


A scatterplot matrix gives a more detailed overview of all pairwise relations.
The plot below suppresses the data points and summarises the relation using
data ellipses and regression lines.
The model syntax, `~ Speed + ... |Dx`, treats `Dx` as a conditioning variable,
giving a separate data ellipse and regression line for each group.

```{r}
#| label: fig-NC-scatmat
#| fig-align: center
#| fig-asp: 1
#| out-width: "100%"
#| fig-cap: "Scatterplot matrix of the `NeuroCog` data. Points are suppressed here, focusing on the data ellipses and regression lines"
scatterplotMatrix(~ Speed + Attention + Memory + Verbal + Visual + ProbSolv | Dx,
  data=NeuroCog,
  plot.points = FALSE,
  smooth = FALSE,
  legend = FALSE,
  col = scales::hue_pal()(3),
  ellipse=list(levels=0.68))
```

**TODO**: Should we add biplot here?

## Fitting the MLM

We proceed to fit the model ...
```{r}
# abbreviate levels for plots
NeuroCog$Dx <- factor(NeuroCog$Dx, 
                      labels = c("Schiz", "SchAff", "Contr"))

NC.mlm <- lm(cbind(Speed, Attention, Memory, Verbal, Visual, ProbSolv) ~ Dx,
             data=NeuroCog)
Anova(NC.mlm)
```

The first research question is captured by the contrasts for the `Dx` factor shown above.
We can test these with `car::linearHypothesis()`.
The contrast `Dx1` for control vs. diagnosed is highly significant, but the second
contrast `Dx2` comparing the schizophrenic and schizoaffective group is not.
```{r NC-linhyp}
# control vs. patients
print(linearHypothesis(NC.mlm, "Dx1"), SSP=FALSE)
```

```{r NC-linhyp2}
# Schizo vs SchizAff
print(linearHypothesis(NC.mlm, "Dx2"), SSP=FALSE)

```

### HE plot

```{r}
#| label: fig-NC-HEplot
#| fig-align: center
#| out-width: "100%"
#| fig-cap: "HE plot of the `NeuroCog` data."
op <- par(mar=c(5,4,1,1)+.1)
heplot(NC.mlm, fill=TRUE, fill.alpha=0.1,
	cex.lab=1.3, cex=1.25)
par(op)
```

