---
title: 'Penguins: MANOVA and HE plots'
author: "Michael Friendly"
date: "6/14/2020"
output: html_document
bibliography:
  - "`r system('kpsewhich graphics.bib', intern=TRUE)`"
  - "`r system('kpsewhich statistics.bib', intern=TRUE)`"
  - "`r system('kpsewhich Rpackages.bib', intern=TRUE)`"
---

```{r setup, include=FALSE}
library(palmerpenguins)
ignore <- suppressMessages(library(dplyr))
# load packages
ignore <- suppressMessages(library(car))
ignore <- suppressMessages(library(rgl))
ignore <- suppressMessages(library(heplots))
ignore <- suppressMessages(library(candisc))
options(digits = 5)
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,   # avoid warnings and messages in the output
  message = FALSE,
  fig.width = 6,
  fig.height = 6,
  tidy.opts=list(width.cutoff = 120),  # For code
  options(width = 90, digits=5) # for output
)
knitr::knit_hooks$set(webgl = hook_webgl)
options(rgl.useNULL = TRUE)
```

## Introduction

The [multivariate EDA](https://rpubs.com/friendly/penguin-biplots) of the `penguins` data brought some simplicity to
understanding the quantitative variables of body mass, flipper length
and the two bill variables. One main virtue of the graphical methods used
there was that `species` could be represented visually in the plot
and annotated with separate data ellipses and regression lines, even though
the principal component analysis and biplot did not include species in the
analysis.

An obvious next step, and the one that brought the `iris` data to prominence is to ask two related multivariate model questions:

* **MANOVA**: To what extent can we be confident that the means of species on these variables reliably differ significantly in multivariate space? A multivariate
analysis of variance (MANOVA) tests group mean differences simultaneously, without appeal to univariate tests or correction for multiple testing.

* **Discriminant analysis**: What weighted sums of the variables best discriminates among the penguin species?  The answer to this question is that the discriminant weights are found in the MANOVA, and are the weights that maximize the multivariate test statistics.

These ideas are implemented in the `heplots` [@heplots] and 
`candisc` [@candisc] packages and described in more detail in other articles [@Friendly:07:manova; @Friendly-etal:ellipses:2013].

### Multivariate visualization

But there is more:

* **HE plots**: Hypothesis-Error plots use ellipses to show the strength of evidence for any multivariate hypothesis

* **Canonical views**: Project the data and ellipses into the low-dimensional space that shows the largest multivariate effects.

### Getting started

```{r}
data(penguins, package="palmerpenguins")
```
As before, we clean the data: abbreviating variable names, making
some character variables factors, and removing missing data.
```{r}
peng <- penguins %>%
	rename(
         bill_length = bill_length_mm, 
         bill_depth = bill_depth_mm, 
         flipper_length = flipper_length_mm, 
         body_mass = body_mass_g
         ) %>%
  mutate(species = as.factor(species),
         island = as.factor(island),
         sex = as.factor(substr(sex,1,1))) %>%
  filter(!is.na(bill_depth),
         !is.na(sex))
```

### Load our packages

```{r load-pkgs, eval=FALSE}
library(car)
library(heplots)
library(candisc)
```

## MANOVA

Let's assume that the goal of an analysis is to determine whether and how the penguin species differ in terms of the four quantitative variables.
A naive multivariate linear model is the following, with species as the only
predictor.

```{r mod0}
peng.mod0 <-lm(cbind(bill_length, bill_depth, flipper_length, body_mass) ~
                 species, data=peng)
Anova(peng.mod0)
```

## HE plots

In a univariate ANOVA, the $F$-test is based on the variability **between** groups relative to the variability **within** groups
$$
F = \frac{SS_H/df_h}{SS_E / df_e}
$$
The analogous multivariate tests are based on the same idea, applied to the between-group sum of squares and cross-products (SSP) matrix, $\mathbf{H}$,
relative to the within-group SSP matrix, $\mathbf{E}$.

To see this visually, consider a plot of the two `bill` beak variables
where we show only the data ellipses for each species and the pooled
data ellipse of the residuals.

```{r covellipses, echo=-1}
op <- par(mar=c(4,4,1,1)+0.1)
cols = c(scales::hue_pal()(3), "black")
covEllipses(peng[,3:4], peng$species, 
            col=cols,
            fill=c(rep(FALSE,3), TRUE), fill.alpha=.1)
```

In the HE plot below,

* the differences **between** species means are shown by what we call the $\mathbf{H}$ ellipsoid, the data ellipsoid
of the fitted (predicted) values under the model.
In 2D, with a two degree-of-freedom test, $\mathbf{H}$ has two dimensions
and appears as a simple ellipse.

* the variation **within** species is reflected in the $\mathbf{E}$ ellipse, which is just 
the pooled data ellipses of the groups, translated to the grand means, or equivalently,
the data ellipse of the residuals.
By default, the size of the ellipse is set to cover 68% of the observations in a bivariate normal sample, an analog of a $\bar{y} \pm 1 s$ univariate
interval. With this 68% interval, you can "read" the residual standard deviation as the half-length of the shadow of the $\mathbf{E}$ ellipse on any axis.
Translating this ellipse to the grand means allows us to show the group centroids on the same scale,
facilitating interpretation.

The orientation of the $\mathbf{H}$ ellipse reflects the negative correlation of the species means:
in general species with larger bill depth have smaller bill length. 

```{r heplot0, echo=-1}
op <- par(mar=c(4,4,1,1)+0.1)
heplot(peng.mod0, fill=TRUE, fill.alpha=0.1, 
       size="effect",
       xlim=c(35,52), ylim=c(14,20))
```


But the overall size of the $\mathbf{H}$ ellipse relative to that of $\mathbf{E}$ is crucial:

* One simple choice, **effect size** scaling, uses $\mathbf{H}/df_e$ to put this on the same scale
as the $\mathbf{E}$ ellipse. This is analogous to an effect size index used in univariate models,
e.g., $ES = (\bar{y}_1 - \bar{y}_2) / s$ in a two-group design.

In the plot above, a visual estimate is that variation of the species means is about ES = 1.5 times that of within-species (residual) variance.  This is a large effect!

* The default used in `heplot` is what we call **significance** (or "evidence") scaling.
That is the relative size of the $\mathbf{H}$ ellipse is set so that in the full p-dimensional
space of all $p$ response variables, the $\mathbf{H}$ ellipse will protrude somewhere
outside the $\mathbf{E}$ ellipse *if and only if* the test for some effect is significant
(by Roy's largest root test)
at a prescribed $\alpha$-level (0.05 by default). This scaling uses $\mathbf{H}/\lambda_\alpha \: df_e$,
where $\lambda_\alpha$ is the critical $\alpha$ value of Roy's test statistic.

```{r heplot1}
heplot(peng.mod0, fill=TRUE, fill.alpha=0.1)
```

The MANOVA test is so overwhelmingly significant that I will stick to effect-size scaling to preserve resolution.

### Contrasts

One lovely virtue of the standard linear model fit with `lm()` is that any ANOVA effect
of $df_h \ge 2$ degrees of freedom can be partitioned into $df_h$ separate 1 df
tests of "linear hypotheses". If these use *orthogonal* contrasts, then the separate $\mathbf{H}$ matrices are additive.

\begin{equation*}
\mathbf{H} = \mathbf{H}_1 + \mathbf{H}_2 + \cdots + \mathbf{H}_{\textrm{df}_h}
\end{equation*}


Let's also assume that in addition to the overall question, the researchers
are interested in two sub-hypotheses:

* Do Gentoo penguins differ from the other two?
* Do Adelie differ from Chinstrap?

These questions can be formulated in terms of two orthogonal contrasts for the species factor. The columns give weights indicating differences between group means.

```{r contrasts}
contrasts(peng$species)<-matrix(c(1,-1,0, -1, -1, -2), 3,2)
contrasts(peng$species)
```

In the `heplots` package, contrasts are specified as a list of tests evaluated by
`car::linearHypothesis`. Each of these generate an $\mathbf{H}$ matrix of rank 1, so these will appear as lines in an HE plot.

```{r refit, echo=-1}
op <- par(mar=c(4,4,1,1)+0.1)
peng.mod0 <-lm(cbind(bill_length, bill_depth, flipper_length, body_mass) ~ species, data=peng)

hyp <- list("A:C"="species1","AC:G"="species2")
heplot(peng.mod0, fill=TRUE, fill.alpha=0.2, 
       hypotheses=hyp, size="effect")
```

This is quite a nice result.  It is clear that Adelie & Chinstrap differ
only in the length of the bill. Gentoo penguins differ from the other two
in having longer, but less deep bill.

### Other plots
`heplot` gives 2D plots for any pair of response variables. For `flipper_length` and `body_mass` we get the plot below.  It is readily seen that these two variables are highly correlated in their species means. We can interpret the major axis of this **H** ellipse as
a measure of pengiun size, with Gentoo being the largest.

```{r heplot34, echo=-1}
op <- par(mar=c(4,4,1,1)+0.1)
heplot(peng.mod0, variables=3:4, 
       size="effect",
       fill=TRUE, fill.alpha=0.2)
```

#### Pairs plots
The `pairs.mlm` method for `mlm` objects gives a all pairwise heplots
in a scatterplot matrix format.
```{r hepairs}
pairs(peng.mod0, size="effect", fill=c(TRUE, FALSE))
```

In these pairwise views, we see some new things:

* The within-group residuals are positively correlated for all pairs. This makes sense to me.

* As we saw above, group means for flipper length and body mass are positively correlated, but

* There are negative correlations of bill depth with the two size variables, and

* Between species bill depth has a correlation $\approx 1$ with body size.  Time to call in the biologists!

#### 3D plots

Finally, there is an `heplot3d()` fundtion that shows the **H** and **E** ellipsoids in 3D space.  This plot should be interactive in an HTML document (drag to rotate, zoom in/out with mouse wheel).

```{r heplot3d, webgl=TRUE}
next3d()
heplot3d(peng.mod0, fill=TRUE, fill.alpha=0.2, size="effect")
```

If you rotate the plot, you will see that the **H** ellipsoid is actually 2-dimensional, because there are $g-1 = 2$ df.

## Other models

A careful researcher would also
wonder, from a modeling perspective, whether differences among species could be accounted for by possible confounders: `island` and `sex`.

```{r mod1}
peng.mod1 <-lm(cbind(bill_length, bill_depth, flipper_length, body_mass) ~
                 species + island + sex, data=peng)
Anova(peng.mod1)
```
`island` seems not to be important, so we delete this

```{r mod2}
peng.mod2 <-lm(cbind(bill_length, bill_depth, flipper_length, body_mass) ~
                 species + sex, data=peng)
Anova(peng.mod2)
```

```{r peng.mod2, echo=-1}
op <- par(mar=c(4,4,1,1)+0.1)
heplot(peng.mod2, size="effect", 
       fill=TRUE, fill.alpha=0.2)
```

So, in addition to the means among species, we also see that male
penguins have both larger and deeper bills.
We could go further to test for an interaction of species with sex, using `cbind( ...) ~ species:sex` for the model.


## Canonical views

The `heplot()`, `pairs.mlm()`, and `heplot3d()` functions provide 2D and 3D views of all effects in a MLM in **variable** space. Alternatively, canonical discriminant analysis (CDA) provides a low-D visualization of between-group variation and vectors reflecting the weights of the response variables on these dimensions in **canonical space**. 

You can imagine this as rotating the 4D space of all four variables,
and projecting the observations on the plane that shows the largest variation in the species means.

In this space, the response variables can be shown as vectors,
whose angles relative to the canonical dimensions reflect their
correlations. The relative lengths of these vectors indicates their contribution to discrimination among species.

This is analogous to a biplot often used in conjunction with a principal component analysis. See: [penguins_biplots](https://rpubs.com/friendly/penguin-biplots)


The printed output of the result of `candisc()` shows that 
70% of the between-species mean differences can be accounted for by one linear combination of the size variables. The second canonical variable accounts for the remaining 13.4%.

```{r}
(peng.can <- candisc(peng.mod0))
```

The `plot()` method for `candisc` objects plots the
canonical discriminant scores in 2D space, with the data ellipses for each group. The variable
vectors reflect the correlations of each variable with the two canonical dimensions.

```{r peng-can-plot, fig.width=8, fig.height=4, out.width="75%", fig.align="center", echo=-1}
op <- par(mar=c(4,4,1,1)+0.1)
col <- scales::hue_pal()(3)
plot(peng.can, col=col, pch=15:17, 
     ellipse=TRUE, scale=9, 
     var.lwd = 2, var.col = "black")
```

From this we can see:

* The first dimension (Can1) separates Gentoo from the other two species. Can2 largely separates Adelie from Chinstrap.

* Can1 is positively correlated with flipper length and body mass,
our penguin "size" variables. But bill depth also separates Gentoo from the others, in the opposite direction 

* Can2 here largely reflects bill length, with Chintrap having longer beaks than the other two.

## References
